---
title: "BD v HC "
author: "Iago"
date: "`r Sys.Date()`"
output: html_document
---


```{r, results='hide'}
source("~/walssbasslab_server/Walss-Bass lab juliet/Iago/1.EWAS_K01_Analysis/Aug.24-EWAS_BDvsHC_BRAINBANK/source_functions/source_functions.R")

# windows path
#source("V:/common/Walss-Bass_Lab/Walss-Bass lab juliet/Iago/1.EWAS_K01_Analysis/Aug.24-EWAS_BDvsHC_BRAINBANK/source_functions/source_functions.R")

need_packages <-c("tibble", "readxl", "dplyr", "EnhancedVolcano", "ggplot2", "ggmanh", "extrafont","gap")#' install by running the function
install_cran_pkgs(pkgs = need_packages)
install_bioconductor_pkgs(pkgs = need_packages)

#' Check if all you need is installed
check_installed(pkgs = need_packages)

#load all packages, if needed
load_pkgs <-c("tibble", "readxl", "dplyr", "EnhancedVolcano", "ggplot2", "ggmanh", "extrafont","gap")
lapply(load_pkgs, require, character.only = TRUE )


```

### 4.1 Read in the data
#### 0.1 Directory variables
```{r paged.print=TRUE; root.dir=TRUE}
#OBS: same paths on FEDORA
inputFolder <- "~/walssbasslab_server/Walss-Bass lab juliet/Iago/1.EWAS_K01_Analysis/Aug.24-EWAS_BDvsHC_BRAINBANK/2.Output_EWAS-models/4.DMPsSheets/"

#OBS: Paths defined to the lab computer
#inputFolder <-  "V:/common/Walss-Bass_Lab/Walss-Bass lab juliet/Iago/1.EWAS_K01_Analysis/Aug.24-EWAS_BDvsHC_BRAINBANK"
outputFolder <- "~/walssbasslab_server/Walss-Bass lab juliet/Iago/1.EWAS_K01_Analysis/Aug.24-EWAS_BDvsHC_BRAINBANK/2.Output_EWAS-models/"
knitr::opts_knit$set(root.dir = inputFolder)
```


#### 0.2 Input variables 
```{r}
date <- "Aug.2024"
# - BD vs. HC DMPs 
DMPs <- read.csv(file.path(inputFolder,"EWAS-BDvsHC_DMPs_Aug.2024.csv"), row.names = 1)

# - BD vs. HC DMPs svaPCs
DMPsSVA <- read.csv(file.path(inputFolder,"EWAS-BDvsHC_DMPs_svaPCs_Aug.2024.csv"), row.names = 1)

# - BDS vs. BDNS DMPs 
DMPsCases <- read.csv(file.path(inputFolder,"EWAS-BDSvsBDNS_DMPs_Aug.2024.csv"), row.names = 1)

# BDs vs. BDNS DMPs sva PCS
DMPsCasesSVA <- read.csv(file.path(inputFolder,"EWAS-BDSvsBDNS_DMPs_svaPCs_Aug.2024.csv"), row.names = 1)
```
#### 0.3 Output variables 
```{r}
rgset_beads_cpgmissing_file <- file.path(inputFolder,"7.Intermediate_files/RGset_goodsamples_no-SNPs_BeadFilter_CpG10perctFilter_BRAIN-BANK_Aug.2024.RData")
```


# BD vs HC Manhatam plot
### NO SVA 
```{r}

DMPs <- read.csv(file.path(inputFolder,"EWAS-BDvsHC_DMPs_Aug.2024.csv"), row.names = 1)
DMPs <- DMPs[which(DMPs$CHR %in% c(1:22)),]
DMPs$CHR <- as.numeric(DMPs$CHR)
DMPs_sorted <- DMPs[order(DMPs$CHR),]
#DMPs_sorted$CHR <- factor(DMPs_sorted$CHR) 


```
#### Defining thresholds
```{r}
# Define two threshold lines
EWSthresh <- -log10(9e-8)


# - Defining the FDR threshold line
p_values <- DMPs_sorted$P.Value

###############################################################################
# Function definition 
FDRthres_function <- function(PVAL,FDRLEVEL){
n <- length(PVAL)

# Sort p-values
sorted_PVAL <- sort(PVAL)

# Calculate BH critical values
bh_critical_values <- (1:n) / n * FDRLEVEL

# Find the largest p-value that is less than or equal to its BH critical value
threshold_index <- max(which(sorted_PVAL <= bh_critical_values))
fdr_threshold <- sorted_PVAL[threshold_index]
return(fdr_threshold)
}
###############################################################################
fdrthres <- FDRthres_function(DMPs_sorted$P.Value,0.05)
FDRthres <- -log10(fdrthres)
```
#### Generating Plot
```{r}

png(filename = file.path(outputFolder,"2.ManhattanPlots/EWAS-BDvsHC_ManhattanPlot_Aug.2024.png"))
g <- DMPs_sorted %>% manhattan_plot(
  .,
  chromosome = NULL,
  signif = c(EWSthresh, FDRthres),
  pval.colname = "P.Value",
  chr.colname = "CHR",
  pos.colname = "MAPINFO",
  rescale = TRUE)

g + 
  scale_color_manual(values = rep(c("gray1","gray"), length.out = 22)) +
  geom_hline(yintercept = EWSthresh, color = "red", linetype = "dashed") +
  geom_hline(yintercept = FDRthres, color = "blue4", linetype = "longdash") +
  ggtitle("BD vs. HC Manhattan Plot") +
  theme(plot.title = element_text(size = 17, face = "bold", family = "Times New Roman"))
dev.off()
```
### With SVA 
```{r}

DMPsSVA <- read.csv(file.path(inputFolder,"EWAS-BDvsHC_DMPs_svaPCs_Aug.2024.csv"), row.names = 1)
DMPsSVA <- DMPsSVA[which(DMPsSVA$CHR %in% c(1:22)),]
DMPsSVA$CHR <- as.numeric(DMPsSVA$CHR)
DMPsSVA_sorted <- DMPsSVA[order(DMPsSVA$CHR),]
#DMPsSVA_sorted$CHR <- factor(DMPsSVA_sorted$CHR) 


```
#### Defining thresholds
```{r}
# Define two threshold lines
EWSthresh <- -log10(9e-8)


# - Defining the FDR threshold line
p_values <- DMPsSVA_sorted$P.Value

###############################################################################
# Function definition 
FDRthres_function <- function(PVAL,FDRLEVEL){
n <- length(PVAL)

# Sort p-values
sorted_PVAL <- sort(PVAL)

# Calculate BH critical values
bh_critical_values <- (1:n) / n * FDRLEVEL

# Find the largest p-value that is less than or equal to its BH critical value
threshold_index <- max(which(sorted_PVAL <= bh_critical_values))
fdr_threshold <- sorted_PVAL[threshold_index]
return(fdr_threshold)
}
###############################################################################
fdrthres <- FDRthres_function(DMPsSVA_sorted$P.Value,0.05)
FDRthres <- -log10(fdrthres)
```
#### Generating Plot
```{r}

png(filename = file.path(outputFolder,"2.ManhattanPlots/EWAS-BDvsHC_ManhattanPlot_svaPCs_Aug.2024.png"))
g <- DMPsSVA_sorted %>% manhattan_plot(
  .,
  chromosome = NULL,
  signif = c(EWSthresh, FDRthres),
  pval.colname = "P.Value",
  chr.colname = "CHR",
  pos.colname = "MAPINFO",
  rescale = TRUE)

g + 
  scale_color_manual(values = rep(c("gray1","gray"), length.out = 22)) +
  geom_hline(yintercept = EWSthresh, color = "red", linetype = "dashed") +
  geom_hline(yintercept = FDRthres, color = "blue4", linetype = "longdash") +
  ggtitle("BD vs. HC Manhattan Plot") +
  theme(plot.title = element_text(size = 17, face = "bold", family = "Times New Roman"))
dev.off()
```
# BD vs. HC QQplot
### NO SVA
```{r}

lambda_gc <- gc.lambda(DMPs_sorted$P.Value)

png(filename = file.path(outputFolder,"1.QQplots/EWAS-BDvsHC_QQPlot_Aug.2024.png"))

qq <- ggmanh::qqunif(x= DMPs_sorted$P.Value)

qq + ggtitle("BD vs. HC QQPlot") +
  theme(plot.title = element_text(size = 17, face = "bold", family = "Times New Roman")) + 
  annotate("text", x = Inf, y = Inf, label = paste("lambda GC =", round(lambda_gc, 3)), 
           hjust = 1.1, vjust = 1.1, size = 5, color = "red")
dev.off()
```

### WITH SVA
```{r}

lambda_gc <- gc.lambda(DMPsSVA_sorted$P.Value)

png(filename = file.path(outputFolder,"1.QQplots/EWAS-BDvsHC_QQPlot_svaPCs_Aug.2024.png"))

qq <- ggmanh::qqunif(x= DMPsSVA_sorted$P.Value)

qq + ggtitle("BD vs. HC QQPlot - svaPCs") +
  theme(plot.title = element_text(size = 17, face = "bold", family = "Times New Roman")) + 
  annotate("text", x = Inf, y = Inf, label = paste("lambda GC =", round(lambda_gc, 3)), 
           hjust = 1.1, vjust = 1.1, size = 5, color = "red")
dev.off()
```

# BD vs. HC Volcano plot
### NO SVA
Volcano plot with [EnhancedVolcano](https://bioconductor.org/packages/release/bioc/vignettes/EnhancedVolcano/inst/doc/EnhancedVolcano.html)
```{r}
################################################################################
# 0.05 nominal threshold
################################################################################
pv = 0.05 # set your p-value
fc = 0.5 # set your fold change

DMPs_volcano <- DMPs[,c(7,10)]

colnames(DMPs_volcano) = c('log2FoldChange', 'pvalue')
library(EnhancedVolcano)
BDvsHC_EnhVolc =EnhancedVolcano(DMPs_volcano,
  lab = NA,
  x = 'log2FoldChange',
  y = 'pvalue',
  pCutoff = pv,
  FCcutoff = fc,
  ylim = c(-0.5,10),
  xlim = c(-1.8,1.8),
  labSize = 2.0)
BDvsHC_EnhVolc
#ggsave(paste0("VolcanoPlot_BD_v_HC_Nominal-p",date,".png"),path = "V:/common/Walss-Bass_Lab/Walss-Bass lab juliet/Iago/1.EWAS_K01_Analysis/Aug.24-EWAS_BDvsHC_BRAINBANK/2.Output_EWAS-models/3.VolcanoPlots", device='png', dpi=300, width = 10, height = 7)
ggsave(paste0("VolcanoPlot_BD_v_HC_Nominal-p",date,".png"),path = file.path(outputFolder,"3.VolcanoPlots"), device='png', dpi=300, width = 10, height = 7)
################################################################################
# FDRcorrected p-value 
################################################################################
pv = fdrthres # set your p-value
fc = 10^1000 # set your fold change
DMPs_volcano <- DMPs[,c(7,10)]

colnames(DMPs_volcano) = c('log2FoldChange', 'pvalue')
library(EnhancedVolcano)
BDvsHC_EnhVolc =EnhancedVolcano(DMPs_volcano,
  lab = NA,
  x = 'log2FoldChange',
  y = 'pvalue',
  pCutoff = pv,
  FCcutoff = fc,
  ylim = c(-0.5,10),
  xlim = c(-1.8,1.8),
  labSize = 2.0,
  title = "BD vs. HC Enhanced Volcano FDR-corrected p-value",)
BDvsHC_EnhVolc
#ggsave(paste0("VolcanoPlot_BD_v_HC_FDR-corrected",date,".png"),path = "V:/common/Walss-Bass_Lab/Walss-Bass lab juliet/Iago/1.EWAS_K01_Analysis/Aug.24-EWAS_BDvsHC_BRAINBANK/2.Output_EWAS-models/3.VolcanoPlots", device='png', dpi=300, width = 10, height = 7)
ggsave(paste0("VolcanoPlot_BD_v_HC_FDR-corrected",date,".png"),path = file.path(outputFolder,"3.VolcanoPlots"), device='png', dpi=300, width = 10, height = 7)

```

### WITH SVA
```{r}
################################################################################
# 0.05 nominal threshold
################################################################################
pv = 0.05 # set your p-value
fc = 0.5 # set your fold change

DMPsSVA_volcano <- DMPsSVA[,c(7,10)]

colnames(DMPsSVA_volcano) = c('log2FoldChange', 'pvalue')
library(EnhancedVolcano)
BDvsHC_EnhVolc =EnhancedVolcano(DMPsSVA_volcano,
  lab = NA,
  x = 'log2FoldChange',
  y = 'pvalue',
  pCutoff = pv,
  FCcutoff = fc,
  ylim = c(-0.5,10),
  xlim = c(-1.8,1.8),
  labSize = 2.0,
  title = "BD vs. HC Enhanced Volcano Nominal p-value - svaPCs")
BDvsHC_EnhVolc
#ggsave(paste0("VolcanoPlot_BD_v_HC_Nominal-p",date,".png"),path = "V:/common/Walss-Bass_Lab/Walss-Bass lab juliet/Iago/1.EWAS_K01_Analysis/Aug.24-EWAS_BDvsHC_BRAINBANK/2.Output_EWAS-models/3.VolcanoPlots", device='png', dpi=300, width = 10, height = 7)
ggsave(paste0("VolcanoPlot_BD_v_HC_Nominal-p_svaPCs",date,".png"),path = file.path(outputFolder,"3.VolcanoPlots"), device='png', dpi=300, width = 10, height = 7)
################################################################################
# FDRcorrected p-value 
################################################################################
pv = fdrthres # set your p-value
fc = 10^1000 # set your fold change
DMPsSVA_volcano <- DMPsSVA[,c(7,10)]

colnames(DMPsSVA_volcano) = c('log2FoldChange', 'pvalue')
library(EnhancedVolcano)
BDvsHC_EnhVolc =EnhancedVolcano(DMPsSVA_volcano,
  lab = NA,
  x = 'log2FoldChange',
  y = 'pvalue',
  pCutoff = pv,
  FCcutoff = fc,
  ylim = c(-0.5,10),
  xlim = c(-1.8,1.8),
  labSize = 2.0,
  title = "BD vs. HC Enhanced Volcano FDR-corrected p-value - svaPCs",)
BDvsHC_EnhVolc
#ggsave(paste0("VolcanoPlot_BD_v_HC_FDR-corrected",date,".png"),path = "V:/common/Walss-Bass_Lab/Walss-Bass lab juliet/Iago/1.EWAS_K01_Analysis/Aug.24-EWAS_BDvsHC_BRAINBANK/2.Output_EWAS-models/3.VolcanoPlots", device='png', dpi=300, width = 10, height = 7)
ggsave(paste0("VolcanoPlot_BD_v_HC_FDR-corrected_svaPCs",date,".png"),path = file.path(outputFolder,"3.VolcanoPlots"), device='png', dpi=300, width = 10, height = 7)

```


***
# BDS vs. BDNS Manhatam plot
### NO SVA
```{R}
DMPsCases <- read.csv(file.path(inputFolder,"EWAS-BDSvsBDNS_DMPs_Aug.2024.csv"), row.names = 1)
DMPsCases <- DMPsCases[which(DMPsCases$CHR %in% c(1:22)),]
DMPsCases$CHR <- as.numeric(DMPsCases$CHR)

DMPsCases_sorted <- DMPsCases[order(DMPsCases$CHR),]

#DMPsCases$CHR <- factor(DMPsCases$CHR) 


```
#### Defining thresholds
```{r}
# Define two threshold lines
EWSthresh <- -log10(9e-8)


# - Defining the FDR threshold line
p_values <- DMPsCases$P.Value

###############################################################################
# Function definition 
FDRthres_function <- function(PVAL,FDRLEVEL){
n <- length(PVAL)

# Sort p-values
sorted_PVAL <- sort(PVAL)

# Calculate BH critical values
bh_critical_values <- (1:n) / n * FDRLEVEL

# Find the largest p-value that is less than or equal to its BH critical value
threshold_index <- max(which(sorted_PVAL <= bh_critical_values))
fdr_threshold <- sorted_PVAL[threshold_index]
return(fdr_threshold)
}
###############################################################################
fdrthres <- FDRthres_function(DMPsCases$P.Value,0.05)
FDRthres <- -log10(fdrthres)
```

#### Generating Plot
```{r}
#font_import()
#loadfonts(device = "win")
png(filename = file.path(outputFolder,"/2.ManhattanPlots/EWAS-BDSvsBDNS_ManhattanPlot_Aug.2024.png"))
g <- DMPsCases_sorted %>% manhattan_plot(
  .,
  chromosome = NULL,
  signif = c(EWSthresh, FDRthres),
  pval.colname = "P.Value",
  chr.colname = "CHR",
  pos.colname = "MAPINFO",
  rescale = TRUE)

g + 
  scale_color_manual(values = rep(c("gray1","gray"), length.out = 22)) +
  geom_hline(yintercept = EWSthresh, color = "red", linetype = "dashed") +
  geom_hline(yintercept = FDRthres, color = "blue4", linetype = "longdash") +
  ggtitle("BDS vs. BDNS Manhattan Plot") +
  theme(plot.title = element_text(size = 17, face = "bold", family = "Times New Roman"))
dev.off()
```
# BDS vs. BDNS Manhatam plot
### WITH SVA
```{R}
DMPsCasesSVA <- read.csv(file.path(inputFolder,"EWAS-BDSvsBDNS_DMPs_svaPCs_Aug.2024.csv"), row.names = 1)
DMPsCasesSVA <- DMPsCasesSVA[which(DMPsCasesSVA$CHR %in% c(1:22)),]
DMPsCasesSVA$CHR <- as.numeric(DMPsCasesSVA$CHR)

DMPsCasesSVA_sorted <- DMPsCasesSVA[order(DMPsCasesSVA$CHR),]

#DMPsCasesSVA$CHR <- factor(DMPsCasesSVA$CHR) 


```
#### Defining thresholds
```{r}
# Define two threshold lines
EWSthresh <- -log10(9e-8)


# - Defining the FDR threshold line
p_values <- DMPsCasesSVA$P.Value

###############################################################################
# Function definition 
FDRthres_function <- function(PVAL,FDRLEVEL){
n <- length(PVAL)

# Sort p-values
sorted_PVAL <- sort(PVAL)

# Calculate BH critical values
bh_critical_values <- (1:n) / n * FDRLEVEL

# Find the largest p-value that is less than or equal to its BH critical value
threshold_index <- max(which(sorted_PVAL <= bh_critical_values))
fdr_threshold <- sorted_PVAL[threshold_index]
return(fdr_threshold)
}
###############################################################################
fdrthres <- FDRthres_function(DMPsCasesSVA$P.Value,0.05)
FDRthres <- -log10(fdrthres)
```

#### Generating Plot
```{r}
#font_import()
#loadfonts(device = "win")
png(filename = file.path(outputFolder,"/2.ManhattanPlots/EWAS-BDSvsBDNS_ManhattanPlot_svaPCs_Aug.2024.png"))
g <- DMPsCasesSVA_sorted %>% manhattan_plot(
  .,
  chromosome = NULL,
  signif = c(EWSthresh, FDRthres),
  pval.colname = "P.Value",
  chr.colname = "CHR",
  pos.colname = "MAPINFO",
  rescale = TRUE)

g + 
  scale_color_manual(values = rep(c("gray1","gray"), length.out = 22)) +
  geom_hline(yintercept = EWSthresh, color = "red", linetype = "dashed") +
  geom_hline(yintercept = FDRthres, color = "blue4", linetype = "longdash") +
  ggtitle("BDS vs. BDNS Manhattan Plot - svaPCs") +
  theme(plot.title = element_text(size = 17, face = "bold", family = "Times New Roman"))
dev.off()
```
# BDS vs. BDNS QQplot
### NO SVA
```{r}

lambda_gc <- gc.lambda(DMPsCases_sorted$P.Value)

png(filename = file.path(outputFolder,"1.QQplots/EWAS-BDSvsBDNS_QQPlot_Aug.2024.png"))

qq <- ggmanh::qqunif(x= DMPsCases_sorted$P.Value)

qq + ggtitle("BDS vs. BDNS QQPlot") +
  theme(plot.title = element_text(size = 17, face = "bold", family = "Times New Roman")) + 
  annotate("text", x = Inf, y = Inf, label = paste("lambda GC =", round(lambda_gc, 3)), 
           hjust = 1.1, vjust = 1.1, size = 5, color = "red")
dev.off()
```
### WITH SVA
```{r}

lambda_gc <- gc.lambda(DMPsCasesSVA_sorted$P.Value)

png(filename = file.path(outputFolder,"1.QQplots/EWAS-BDSvsBDNS_QQPlot_svaPCs_Aug.2024.png"))

qq <- ggmanh::qqunif(x= DMPsCasesSVA_sorted$P.Value)

qq + ggtitle("BDS vs. BDNS QQPlot") +
  theme(plot.title = element_text(size = 17, face = "bold", family = "Times New Roman")) + 
  annotate("text", x = Inf, y = Inf, label = paste("lambda GC =", round(lambda_gc, 3)), 
           hjust = 1.1, vjust = 1.1, size = 5, color = "red")
dev.off()
```

# BDS vs. BDNS Volcano Plot
### NO SVA
```{r}
################################################################################
# 0.05 nominal threshold
################################################################################
pv = 0.05 # set your p-value
fc = 0.5 # set your fold change

DMPs_volcano <- DMPsCases[,c(7,10)]

colnames(DMPs_volcano) = c('log2FoldChange', 'pvalue')
library(EnhancedVolcano)
BDSvsBDNS_EnhVolc =EnhancedVolcano(DMPs_volcano,
  lab = NA,
  x = 'log2FoldChange',
  y = 'pvalue',
  pCutoff = pv,
  FCcutoff = fc,
  ylim = c(-0.5,10),
  xlim = c(-1.8,1.8),
  labSize = 2.0,
    title = "BDS vs. BDNS Enhanced Volcano Nominal p-value",)+
  theme(plot.title = element_text(size = 15))
BDSvsBDNS_EnhVolc
ggsave(paste0("VolcanoPlot_BDS_v_BDNS_Nominal-p",date,".png"),path = file.path(outputFolder,"3.VolcanoPlots"), device='png', dpi=300, width = 10, height = 7)
################################################################################
# FDRcorrected p-value 
################################################################################
pv = fdrthres # set your p-value
fc = 10^1000 # set your fold change
DMPs_volcano <- DMPsCases[,c(7,10)]

colnames(DMPs_volcano) = c('log2FoldChange', 'pvalue')
library(EnhancedVolcano)
BDSvsBDNS_EnhVolc =EnhancedVolcano(DMPs_volcano,
  lab = NA,
  x = 'log2FoldChange',
  y = 'pvalue',
  pCutoff = pv,
  FCcutoff = fc,
  ylim = c(-0.5,10),
  xlim = c(-1.8,1.8),
  labSize = 1.0,
  title = "BDS vs. BDNS Enhanced Volcano FDR-corrected p-value",) +
  theme(plot.title = element_text(size = 15))
BDSvsBDNS_EnhVolc
ggsave(paste0("VolcanoPlot_BDS_v_BDNS_FDR-corrected",date,".png"),path =file.path(outputFolder,"3.VolcanoPlots"), device='png', dpi=300, width = 10, height = 7)

```

### WITH SVA
```{r}
################################################################################
# 0.05 nominal threshold
################################################################################
pv = 0.05 # set your p-value
fc = 0.5 # set your fold change

DMPs_volcano <- DMPsCasesSVA[,c(7,10)]

colnames(DMPs_volcano) = c('log2FoldChange', 'pvalue')
library(EnhancedVolcano)
BDSvsBDNS_EnhVolc =EnhancedVolcano(DMPs_volcano,
  lab = NA,
  x = 'log2FoldChange',
  y = 'pvalue',
  pCutoff = pv,
  FCcutoff = fc,
  ylim = c(-0.5,10),
  xlim = c(-1.8,1.8),
  labSize = 2.0,
    title = "BDS vs. BDNS Enhanced Volcano Nominal p-value sva PCs",)+
  theme(plot.title = element_text(size = 12))
BDSvsBDNS_EnhVolc
ggsave(paste0("VolcanoPlot_BDS_v_BDNS_Nominal-p_svaPCs",date,".png"),path = file.path(outputFolder,"3.VolcanoPlots"), device='png', dpi=300, width = 10, height = 7)
################################################################################
# FDRcorrected p-value 
################################################################################
pv = fdrthres # set your p-value
fc = 10^1000 # set your fold change
DMPs_volcano <- DMPsCasesSVA[,c(7,10)]

colnames(DMPs_volcano) = c('log2FoldChange', 'pvalue')
library(EnhancedVolcano)
BDSvsBDNS_EnhVolc =EnhancedVolcano(DMPs_volcano,
  lab = NA,
  x = 'log2FoldChange',
  y = 'pvalue',
  pCutoff = pv,
  FCcutoff = fc,
  ylim = c(-0.5,10),
  xlim = c(-1.8,1.8),
  labSize = 1.0,
  title = "BDS vs. BDNS Enhanced Volcano FDR-corrected p-value sva PCs",) +
  theme(plot.title = element_text(size = 12))
BDSvsBDNS_EnhVolc
ggsave(paste0("VolcanoPlot_BDS_v_BDNS_FDR-corrected_svaPCs",date,".png"),path =file.path(outputFolder,"3.VolcanoPlots"), device='png', dpi=300, width = 10, height = 7)

```

# BDS vs. HC Volcano Plot
### NO SVA
```{r}
################################################################################
# 0.05 nominal threshold
################################################################################
pv = 0.05 # set your p-value
fc = 0.5 # set your fold change

DMPs_volcano <- DMPsCases[,c(7,10)]

colnames(DMPs_volcano) = c('log2FoldChange', 'pvalue')
library(EnhancedVolcano)
BDSvsHC_EnhVolc =EnhancedVolcano(DMPs_volcano,
  lab = NA,
  x = 'log2FoldChange',
  y = 'pvalue',
  pCutoff = pv,
  FCcutoff = fc,
  ylim = c(-0.5,10),
  xlim = c(-1.8,1.8),
  labSize = 2.0,
    title = "BDS vs. BDNS Enhanced Volcano Nominal p-value",)+
  theme(plot.title = element_text(size = 15))
BDSvsHC_EnhVolc
ggsave(paste0("VolcanoPlot_BDS_v_BDNS_Nominal-p",date,".png"),path = file.path(outputFolder,"3.VolcanoPlots"), device='png', dpi=300, width = 10, height = 7)
################################################################################
# FDRcorrected p-value 
################################################################################
pv = fdrthres # set your p-value
fc = 10^1000 # set your fold change
DMPs_volcano <- DMPsCases[,c(7,10)]

colnames(DMPs_volcano) = c('log2FoldChange', 'pvalue')
library(EnhancedVolcano)
BDSvsHC_EnhVolc =EnhancedVolcano(DMPs_volcano,
  lab = NA,
  x = 'log2FoldChange',
  y = 'pvalue',
  pCutoff = pv,
  FCcutoff = fc,
  ylim = c(-0.5,10),
  xlim = c(-1.8,1.8),
  labSize = 1.0,
  title = "BDS vs. BDNS Enhanced Volcano FDR-corrected p-value",) +
  theme(plot.title = element_text(size = 15))
BDSvsHC_EnhVolc
ggsave(paste0("VolcanoPlot_BDS_v_BDNS_FDR-corrected",date,".png"),path =file.path(outputFolder,"3.VolcanoPlots"), device='png', dpi=300, width = 10, height = 7)

```

### WITH SVA
```{r}
################################################################################
# 0.05 nominal threshold
################################################################################
pv = 0.05 # set your p-value
fc = 0.5 # set your fold change

DMPs_volcano <- DMPsCasesSVA[,c(7,10)]

colnames(DMPs_volcano) = c('log2FoldChange', 'pvalue')
library(EnhancedVolcano)
BDSvsHC_EnhVolc =EnhancedVolcano(DMPs_volcano,
  lab = NA,
  x = 'log2FoldChange',
  y = 'pvalue',
  pCutoff = pv,
  FCcutoff = fc,
  ylim = c(-0.5,10),
  xlim = c(-1.8,1.8),
  labSize = 2.0,
    title = "BDS vs. BDNS Enhanced Volcano Nominal p-value sva PCs",)+
  theme(plot.title = element_text(size = 12))
BDSvsHC_EnhVolc
ggsave(paste0("VolcanoPlot_BDS_v_BDNS_Nominal-p_svaPCs",date,".png"),path = file.path(outputFolder,"3.VolcanoPlots"), device='png', dpi=300, width = 10, height = 7)
################################################################################
# FDRcorrected p-value 
################################################################################
pv = fdrthres # set your p-value
fc = 10^1000 # set your fold change
DMPs_volcano <- DMPsCasesSVA[,c(7,10)]

colnames(DMPs_volcano) = c('log2FoldChange', 'pvalue')
library(EnhancedVolcano)
BDSvsHC_EnhVolc =EnhancedVolcano(DMPs_volcano,
  lab = NA,
  x = 'log2FoldChange',
  y = 'pvalue',
  pCutoff = pv,
  FCcutoff = fc,
  ylim = c(-0.5,10),
  xlim = c(-1.8,1.8),
  labSize = 1.0,
  title = "BDS vs. BDNS Enhanced Volcano FDR-corrected p-value sva PCs",) +
  theme(plot.title = element_text(size = 12))
BDSvsHC_EnhVolc
ggsave(paste0("VolcanoPlot_BDS_v_BDNS_FDR-corrected_svaPCs",date,".png"),path =file.path(outputFolder,"3.VolcanoPlots"), device='png', dpi=300, width = 10, height = 7)

```