---
title: " EWAS - limma 4 DMP Analysis"
author: "Santos, Iago & De La Garza, Stephen"
date: "`r Sys.Date()`"
output: html_document
---

```{r Loading Needed packages}
source("~/walssbasslab_server/Walss-Bass lab juliet/Iago/1.EWAS_K01_Analysis/Aug.24-EWAS_BDvsHC_BRAINBANK/source_functions/source_functions.R")

need_packages <-c("tibble", "readxl", "dplyr", "limma", "edgeR", "clusterProfiler", "enrichplot", "ggplot2")#' install by running the function
install_cran_pkgs(pkgs = need_packages)
install_bioconductor_pkgs(pkgs = need_packages)

#' Check if all you need is installed
check_installed(pkgs = need_packages)

#load all packages, if needed
load_pkgs <- c("tibble", "readxl", "dplyr", "limma", "edgeR", "clusterProfiler", "enrichplot", "ggplot2")
lapply(load_pkgs, require, character.only = TRUE )
```

# Read in the data

#### 0.1 Directory variables

```{r paged.print=TRUE, root.dir=TRUE}
inputFolder <-  "/home/frieslab/walssbasslab_server/Walss-Bass lab juliet/Iago/1.EWAS_K01_Analysis/Aug.24-EWAS_BDvsHC_BRAINBANK"
outputFolder <- inputFolder
knitr::opts_knit$set(root.dir = inputFolder)
```

#### 0.2 Input variables

```{r}
date <- "Aug.2024"
# - QN Mvalues
load(file.path(inputFolder, "7.Intermediate_files/1.Betas/QNMvalues_BRAINBANK_Aug.2024.RData"))

# - Meta data
meta <- read.csv(file.path(inputFolder, "6.Phenotype_files/K01_annotation_QCmetrics_removalAug.2024.csv"))

# - Neuronal Proportion Meta data

metaNP <- read_excel(file.path(inputFolder,"6.Phenotype_files/1.NeuronalProportion/K01_neuronalproportion_9.17.24.xlsx"))

# - manifest
manifest <- file.path(inputFolder,"infinium-methylationepic-v-1-0-b5-manifest-file.csv")

```

#### 0.3 Output variables


------------------------------------------------------------------------

# 4.1 Preparing data for regression ----

#### Merging Neuronal Proportion meta data with full metadata ----

\#### OPITINAL \####

If your metaNP data has barcode column named "Barcode", you may change to "barcode" to match the full meta data. That way, we can use inner_join()

```{r}

metaNP <- metaNP %>% rename("barcode" = "Barcode")
meta <- meta %>% rename("Brain_Bank" = "Brain_.Bank")

meta <- inner_join(x =meta, y = metaNP, by = "barcode")
```

#### Changing Brain_Bank names to readable names ----

It will be necessary in this case. Some Brain bank names have **"\\s"** or **" \* "** characters, that aren't valid naming characters to R.

```{r Changing Brain Bank in order to remove whitespaces and special characters}
BB <- meta$Brain_Bank
pattern = "[ *]"
BB <- gsub(pattern = pattern, replacement = "_", x = BB)
meta$Brain_Bank <- BB

```

# Converting Labels into information ----

#### Creating different groups 
- Case/control comparison
- BD vs BDNS comparison 

Obs: From these, we can create BDSvc.HC and BNDSvs.HC
```{r}

# - Case control column 
meta$Group_diagnosis <- ifelse(meta$Group == 3, "HC", "BD")
table(meta$Group_diagnosis)


# - BDS vs BDNS

group_labels <- character(length(meta$Group))

for (i in 1:length(meta$Group)) {
  if (meta$Group[i] == 2) {
    group_labels[i] <- "BDS"
  } else if (meta$Group[i] == 1) {
    group_labels[i] <- "BDNS"
  } else if (meta$Group[i] == 3) {
    group_labels[i] <- "HC"
  }
}

meta$GroupBDSvsBDNS <- group_labels

table(meta$GroupBDSvsBDNS)

```

#### Removing duplicates based on barcode
```{r}
load(file.path(inputFolder,"7.Intermediate_files/Extra-duplicates/DUPLICATES_TO_REMOVE.RData"))

meta <- meta[which(!meta$barcode %in% targets_remove_barcode) ,]
dim(meta)
```


```{r}
write.csv(meta, file = file.path(inputFolder, "6.Phenotype_files/K01_annotation_QCmetrics_removal_EWASready_Aug.2024.csv"), row.names = F,)
```



# 4.3 making models

### 4.3.1) BD *vs.* HC
```{r}
meta <- read.csv(file.path(inputFolder, "6.Phenotype_files/K01_annotation_QCmetrics_removal_EWASready_Aug.2024.csv"))

mantein <- colnames(meta)[colnames(meta)%in% c("Group_diagnosis","barcode","Age","Sex","PMI","Brain_Bank","Neuronal_prop", "Group", "GroupBDSvsBDNS","svaPC1","svaPC2","svaPC3","svaPC4")]
                          
print(mantein)

meta <- meta[,colnames(meta) %in% mantein]

factors <- c("Sex","Brain_Bank","barcode","Group_diagnosis", "Group", "GroupBDSvsBDNS")
for (cols in factors){
  meta[,cols] <- as.factor(meta[,cols])
}

numerics <- colnames(meta)[!names(meta) %in% c(factors)]
for (cols in numerics){
  meta[,cols] <- as.numeric(meta[,cols])
}

str(meta)
```

Create the group names that you want to compare. In this case, we are comparing BD to HC.

```{r}
group = meta$Group_diagnosis
age = meta$Age
Brain_Bank = meta$Brain_Bank
Sex = meta$Sex
PMI = meta$PMI
Neuronal_prop = meta$Neuronal_prop
svaPC1 = meta$svaPC1
svaPC2 = meta$svaPC2
svaPC3 = meta$svaPC3
svaPC4 = meta$svaPC4

```

***
## a) With svaPC design
```{R}

# use the above to create a design matrix
design <- model.matrix(~0+group+ Age + Sex + PMI + Neuronal_prop +svaPC1 + svaPC2 + svaPC3  + svaPC4, data=meta)
#colnames(design) <- c(levels(Group_diagnosis),levels(Sex)[-1])
 
# fit the linear model 
fit <- lmFit(QNMValues, design)
# create a contrast matrix for specific comparisons
contMatrix <- makeContrasts(groupBD - groupHC,
                           levels=design)
contMatrix
```

#### Fitting the contrasts and summurizing results
How many were upregulated and how many downregulated?
```{r}
# fit the contrasts
fit2 <- contrasts.fit(fit, contMatrix)
fit2 <- eBayes(fit2)

# look at the numbers of DM CpGs at FDR < 0.05
summary(decideTests(fit2))
```

#### Annotating the results using illumina manifest
```{r}

anno <- read.csv(manifest,as.is=TRUE, skip = 7)
annoEPIC <- anno[,c("CHR","MAPINFO",'Name',"UCSC_RefGene_Name","UCSC_RefGene_Accession","UCSC_RefGene_Group")]

annoEPICSub <- annoEPIC[match(rownames(QNMValues),annoEPIC$Name),]

DMPs <- topTable(fit2, num=Inf, coef=1,genelist=annoEPICSub)
head(DMPs)


write.csv(DMPs, file = file.path(inputFolder,"/2.Output_EWAS-models/4.DMPsSheets/1.Base-Model/EWAS-BB-BDvsHC_DMPs_svaPCs_Aug.2024.csv"))

```

***
## b) Whithut sva design
```{R}

# use the above to create a design matrix
design <- model.matrix(~0+group+ Age + Sex + PMI + Neuronal_prop, data=meta)
#colnames(design) <- c(levels(Group_diagnosis),levels(Sex)[-1])
 
# fit the linear model 
fit <- lmFit(QNMValues, design)
# create a contrast matrix for specific comparisons
contMatrix <- makeContrasts(groupBD - groupHC,
                           levels=design)
contMatrix
```

#### Fitting the contrasts and summurizing results
```{r}
# fit the contrasts
fit2 <- contrasts.fit(fit, contMatrix)
fit2 <- eBayes(fit2)

# look at the numbers of DM CpGs at FDR < 0.05
summary(decideTests(fit2))
```
#### Annotating the results using illumina manifest
```{r}

#anno <- read.csv(manifest,as.is=TRUE, skip = 7)
annoEPIC <- anno[,c("CHR","MAPINFO",'Name',"UCSC_RefGene_Name","UCSC_RefGene_Accession","UCSC_RefGene_Group")]

annoEPICSub <- annoEPIC[match(rownames(QNMValues),annoEPIC$Name),]

DMPs <- topTable(fit2, num=Inf, coef=1,genelist=annoEPICSub)
head(DMPs)


write.csv(DMPs, file = file.path(inputFolder,"/2.Output_EWAS-models/4.DMPsSheets/1.Base-Model/EWAS-BB-BDvsHC_DMPs_Aug.2024.csv"))

```

***
### 4.3.2 BDS vc BDNS
#### Reading the meta-data again 
  - Subsetting meta data column
```{R}
meta <- read.csv(file.path(inputFolder, "6.Phenotype_files/K01_annotation_QCmetrics_removal_EWASready_Aug.2024.csv"))

metaCases <- meta[which(meta$Group %in% c(1,2)),]
mantein <- colnames(metaCases)[colnames(metaCases)%in% c("Group_diagnosis","barcode","Age","Sex","PMI","Brain_Bank","Neuronal_prop", "Group", "GroupBDSvsBDNS","svaPC1","svaPC2","svaPC3","svaPC4")]
                          
print(mantein)

metaCases <- metaCases[,colnames(metaCases) %in% mantein]

factors <- c("Sex","Brain_Bank","barcode","Group_diagnosis", "Group", "GroupBDSvsBDNS")
for (cols in factors){
  metaCases[,cols] <- as.factor(metaCases[,cols])
}

numerics <- colnames(metaCases)[!names(metaCases) %in% c(factors)]
for (cols in numerics){
  metaCases[,cols] <- as.numeric(metaCases[,cols])
}

str(metaCases)
table(metaCases$GroupBDSvsBDNS)
```

#### Subsetting QNM-values (Quantile Normalized M-Values) to the individuals we are analysing
```{r}
QNMValuesCases <- QNMValues[,which(colnames(QNMValues) %in% metaCases$barcode)]
dim(QNMValuesCases)
#save(QNMValuesCases, file = file.path(inputFolder,paste0("7.Intermediate_files/1.Betas/QNMMvaluesCases_BRAINBANK_",date,".RData")))
```

#### Creating BDSvsBDNS groups
```{r}

group = metaCases$GroupBDSvsBDNS
Age = metaCases$Age
Brain_Bank = metaCases$Brain_Bank
Sex = metaCases$Sex
PMI = metaCases$PMI
Neuronal_prop = metaCases$Neuronal_prop
svaPC1 = metaCases$svaPC1
svaPC2 = metaCases$svaPC2
svaPC3 = metaCases$svaPC3
svaPC4 = metaCases$svaPC4
```

***
## a) With sva design
```{r}
# use the above to create a design matrix
designCases <- model.matrix(~0+group+ Age + Sex + PMI + Neuronal_prop + svaPC1 + svaPC2 + svaPC3 + svaPC4, data=metaCases)
#colnames(design) <- c(levels(Group_diagnosis),levels(Sex)[-1])
 
# fit the linear model 
fit <- lmFit(QNMValuesCases, designCases)
# create a contrast matrix for specific comparisons
contMatrix <- makeContrasts(groupBDS - groupBDNS,
                           levels=designCases)
contMatrix
```

```{r}
# fit the contrasts
fit2 <- contrasts.fit(fit, contMatrix)
fit2 <- eBayes(fit2)

# look at the numbers of DM CpGs at FDR < 0.05
summary(decideTests(fit2))
```

```{r}

anno <- read.csv(manifest,as.is=TRUE, skip = 7)
annoEPIC <- anno[,c("CHR","MAPINFO",'Name',"UCSC_RefGene_Name","UCSC_RefGene_Accession","UCSC_RefGene_Group")]

annoEPICSub <- annoEPIC[match(rownames(QNMValues),annoEPIC$Name),]

DMPsCases <- topTable(fit2, num=Inf, coef=1,genelist=annoEPICSub)
head(DMPsCases)


write.csv(DMPsCases, file = file.path(inputFolder,"2.Output_EWAS-models/4.DMPsSheets/1.Base-Model/EWAS-BB-BDSvsBDNS_DMPs_svaPCs_Aug.2024.csv"))

```

***
## b) Without sva design
```{r}
# use the above to create a design matrix
designCases <- model.matrix(~0+group+ Age + Sex + PMI + Neuronal_prop, data=metaCases)
#colnames(design) <- c(levels(Group_diagnosis),levels(Sex)[-1])
 
# fit the linear model 
fit <- lmFit(QNMValuesCases, designCases)
# create a contrast matrix for specific comparisons
contMatrix <- makeContrasts(groupBDS - groupBDNS,
                           levels=designCases)
contMatrix
```

```{r}
# fit the contrasts
fit2 <- contrasts.fit(fit, contMatrix)
fit2 <- eBayes(fit2)

# look at the numbers of DM CpGs at FDR < 0.05
summary(decideTests(fit2))
```

```{r}

# anno <- read.csv(manifest,as.is=TRUE, skip = 7)
annoEPIC <- anno[,c("CHR","MAPINFO",'Name',"UCSC_RefGene_Name","UCSC_RefGene_Accession","UCSC_RefGene_Group")]

annoEPICSub <- annoEPIC[match(rownames(QNMValues),annoEPIC$Name),]

DMPsCases <- topTable(fit2, num=Inf, coef=1,genelist=annoEPICSub)
head(DMPsCases)


write.csv(DMPsCases, file = file.path(inputFolder,"2.Output_EWAS-models/4.DMPsSheets/1.Base-Model/EWAS-BB-BDSvsBDNS_DMPs_Aug.2024.csv"))

```

***
### 4.3.3 HC *vs.* BDS
#### Reading the meta-data again 
  - Subsetting meta data column
```{r}
meta <- read.csv(file.path(inputFolder, "6.Phenotype_files/K01_annotation_QCmetrics_removal_EWASready_Aug.2024.csv"))

metaHC_BDS <- meta[which(meta$Group %in% c(2,3)),]
mantein <- colnames(metaHC_BDS)[colnames(metaHC_BDS)%in% c("Group_diagnosis","barcode","Age","Sex","PMI","Brain_Bank","Neuronal_prop", "Group", "GroupBDSvsBDNS","svaPC1","svaPC2","svaPC3","svaPC4")]
                          
print(mantein)

metaHC_BDS <- metaHC_BDS[,colnames(metaHC_BDS) %in% mantein]

factors <- c("Sex","Brain_Bank","barcode","Group_diagnosis", "Group", "GroupBDSvsBDNS")
for (cols in factors){
  metaHC_BDS[,cols] <- as.factor(metaHC_BDS[,cols])
}

numerics <- colnames(metaHC_BDS)[!names(metaHC_BDS) %in% c(factors)]
for (cols in numerics){
  metaHC_BDS[,cols] <- as.numeric(metaHC_BDS[,cols])
}

str(metaHC_BDS)
```


#### Subsetting QNM-values (Quantile Normalized M-Values) to the individuals we are analysing
```{r}
QNMValuesHC_BDS <- QNMValues[,which(colnames(QNMValues) %in% metaHC_BDS$barcode)]
```

Create the group names that you want to compare. In this case, we are comparing BD to HC.

```{r}
group = metaHC_BDS$GroupBDSvsBDNS
Age = metaHC_BDS$Age
Brain_Bank = metaHC_BDS$Brain_Bank
Sex = metaHC_BDS$Sex
PMI = metaHC_BDS$PMI
Neuronal_prop = metaHC_BDS$Neuronal_prop
svaPC1 = metaHC_BDS$svaPC1
svaPC2 = metaHC_BDS$svaPC2
svaPC3 = metaHC_BDS$svaPC3
svaPC4 = metaHC_BDS$svaPC4

```

#### With SVA design 
```{R}
# use the above to create a design matrix
design <- model.matrix(~0+group + Age + Sex + PMI + Neuronal_prop +svaPC1 + svaPC2 + svaPC3  + svaPC4, data=metaHC_BDS)
#colnames(design) <- c(levels(Group_diagnosis),levels(Sex)[-1])
 
# fit the linear model 
fit <- lmFit(QNMValuesHC_BDS, design)
# create a contrast matrix for specific comparisons
contMatrix <- makeContrasts(groupBDS - groupHC,
                           levels=design)
contMatrix
```

```{r}
# fit the contrasts
fit2 <- contrasts.fit(fit, contMatrix)
fit2 <- eBayes(fit2)

# look at the numbers of DM CpGs at FDR < 0.05
summary(decideTests(fit2))
```

```{r}

# anno <- read.csv(manifest,as.is=TRUE, skip = 7)
annoEPIC <- anno[,c("CHR","MAPINFO",'Name',"UCSC_RefGene_Name","UCSC_RefGene_Accession","UCSC_RefGene_Group")]

annoEPICSub <- annoEPIC[match(rownames(QNMValues),annoEPIC$Name),]

DMPsHC_BDS <- topTable(fit2, num=Inf, coef=1,genelist=annoEPICSub)
head(DMPsHC_BDS)


write.csv(DMPsHC_BDS, file = file.path(inputFolder,"/2.Output_EWAS-models/4.DMPsSheets/1.Base-Model/EWAS-BB-BDSvsHC_DMPs_svaPCs_Aug.2024.csv"))

```

#### without SVA design 
```{r}
# use the above to create a design matrix
design <- model.matrix(~0+group+ Age + Sex + PMI + Neuronal_prop, data=metaHC_BDS)
#colnames(design) <- c(levels(Group_diagnosis),levels(Sex)[-1])
 
# fit the linear model 
fit <- lmFit(QNMValuesHC_BDS, design)
# create a contrast matrix for specific comparisons
contMatrix <- makeContrasts(groupBDS - groupHC,
                           levels=design)
contMatrix
```

```{r}
# fit the contrasts
fit2 <- contrasts.fit(fit, contMatrix)
fit2 <- eBayes(fit2)

# look at the numbers of DM CpGs at FDR < 0.05
summary(decideTests(fit2))
```

```{r}

# anno <- read.csv(manifest,as.is=TRUE, skip = 7)
annoEPIC <- anno[,c("CHR","MAPINFO",'Name',"UCSC_RefGene_Name","UCSC_RefGene_Accession","UCSC_RefGene_Group")]

annoEPICSub <- annoEPIC[match(rownames(QNMValues),annoEPIC$Name),]

DMPsHC_BDS  <- topTable(fit2, num=Inf, coef=1,genelist=annoEPICSub)
head(DMPsHC_BDS )


write.csv(DMPsHC_BDS , file = file.path(inputFolder,"2.Output_EWAS-models/4.DMPsSheets/1.Base-Model/EWAS-BB-BDSvsHC_DMPs_Aug.2024.csv"))

```

***
### 4.3.4  BDNS  *vs.*HC
```{r}
meta <- read.csv(file.path(inputFolder, "6.Phenotype_files/K01_annotation_QCmetrics_removal_EWASready_Aug.2024.csv"))

# Manteining groups 1) BDNS and 3) HC
metaHC_BDNS <- meta[which(meta$Group %in% c(1,3)),]
mantein <- colnames(metaHC_BDNS)[colnames(metaHC_BDNS)%in% c("Group_diagnosis","barcode","Age","Sex","PMI","Brain_Bank","Neuronal_prop", "Group", "GroupBDSvsBDNS","svaPC1","svaPC2","svaPC3","svaPC4")]
                          
print(mantein)

metaHC_BDNS <- metaHC_BDNS[,colnames(metaHC_BDNS) %in% mantein]

factors <- c("Sex","Brain_Bank","barcode","Group_diagnosis", "Group", "GroupBDSvsBDNS")
for (cols in factors){
  metaHC_BDNS[,cols] <- as.factor(metaHC_BDNS[,cols])
}

numerics <- colnames(metaHC_BDNS)[!names(metaHC_BDNS) %in% c(factors)]
for (cols in numerics){
  metaHC_BDNS[,cols] <- as.numeric(metaHC_BDNS[,cols])
}

str(metaHC_BDNS)
```


Filtering QNMvalues to fit specific meta 
```{r}
QNMValuesHC_BDNS <- QNMValues[,which(colnames(QNMValues) %in% metaHC_BDNS$barcode)]
```

Create the group names that you want to compare. In this case, we are comparing BD to HC.

```{r}
group = metaHC_BDNS$GroupBDSvsBDNS
Age = metaHC_BDNS$Age
Brain_Bank = metaHC_BDNS$Brain_Bank
Sex = metaHC_BDNS$Sex
PMI = metaHC_BDNS$PMI
Neuronal_prop = metaHC_BDNS$Neuronal_prop
svaPC1 = metaHC_BDNS$svaPC1
svaPC2 = metaHC_BDNS$svaPC2
svaPC3 = metaHC_BDNS$svaPC3
svaPC4 = metaHC_BDNS$svaPC4

```

```{R}
# use the above to create a design matrix
design <- model.matrix(~0+group + Age + Sex + PMI + Neuronal_prop +svaPC1 + svaPC2 + svaPC3  + svaPC4, data=metaHC_BDNS)
#colnames(design) <- c(levels(Group_diagnosis),levels(Sex)[-1])
 
# fit the linear model 
fit <- lmFit(QNMValuesHC_BDNS, design)
# create a contrast matrix for specific comparisons
contMatrix <- makeContrasts(groupBDNS - groupHC,
                           levels=design)
contMatrix
```

```{r}
# fit the contrasts
fit2 <- contrasts.fit(fit, contMatrix)
fit2 <- eBayes(fit2)

# look at the numbers of DM CpGs at FDR < 0.05
summary(decideTests(fit2))
```

```{r}

anno <- read.csv(manifest,as.is=TRUE, skip = 7)
annoEPIC <- anno[,c("CHR","MAPINFO",'Name',"UCSC_RefGene_Name","UCSC_RefGene_Accession","UCSC_RefGene_Group")]

annoEPICSub <- annoEPIC[match(rownames(QNMValues),annoEPIC$Name),]

DMPsHC_BDNS <- topTable(fit2, num=Inf, coef=1,genelist=annoEPICSub)
head(DMPsHC_BDNS)


write.csv(DMPsHC_BDNS, file = file.path(inputFolder,"/2.Output_EWAS-models/4.DMPsSheets/1.Base-Model/EWAS-BB-BDNSvsHC_DMPs_SVA_Aug.2024.csv"))

```

#### without SVA design 
```{r}
# use the above to create a design matrix
design <- model.matrix(~0+group+ Age + Sex + PMI + Neuronal_prop, data=metaHC_BDNS)
#colnames(design) <- c(levels(Group_diagnosis),levels(Sex)[-1])
 
# fit the linear model 
fit <- lmFit(QNMValuesHC_BDNS, design)
# create a contrast matrix for specific comparisons
contMatrix <- makeContrasts(groupBDNS - groupHC,
                           levels=design)
contMatrix
```

```{r}
# fit the contrasts
fit2 <- contrasts.fit(fit, contMatrix)
fit2 <- eBayes(fit2)

# look at the numbers of DM CpGs at FDR < 0.05
summary(decideTests(fit2))
```

```{r}

# anno <- read.csv(manifest,as.is=TRUE, skip = 7)
annoEPIC <- anno[,c("CHR","MAPINFO",'Name',"UCSC_RefGene_Name","UCSC_RefGene_Accession","UCSC_RefGene_Group")]

annoEPICSub <- annoEPIC[match(rownames(QNMValues),annoEPIC$Name),]

DMPsHC_BDNS  <- topTable(fit2, num=Inf, coef=1,genelist=annoEPICSub)
head(DMPsHC_BDNS )


write.csv(DMPsHC_BDNS , file = file.path(inputFolder,"2.Output_EWAS-models/4.DMPsSheets/1.Base-Model/EWAS-BB-BDNSvsHC_DMPs_Aug.2024.csv"))

```
