---
title: "BDNS vs BDS "
author: "Iago"
date: "`r Sys.Date()`"
output: html_document
---


```{r, results='hide'}
source("~/walssbasslab_server/Walss-Bass lab juliet/Iago/1.EWAS_K01_Analysis/Aug.24-EWAS_BDvsHC_BRAINBANK/source_functions/source_functions.R")

# windows path
#source("V:/common/Walss-Bass_Lab/Walss-Bass lab juliet/Iago/1.EWAS_K01_Analysis/Aug.24-EWAS_BDNSvsBDS_BRAINBANK/source_functions/source_functions.R")

need_packages <-c("tibble", "readxl", "dplyr", "EnhancedVolcano", "ggplot2", "ggmanh", "extrafont","gap")#' install by running the function
install_cran_pkgs(pkgs = need_packages)
install_bioconductor_pkgs(pkgs = need_packages)

#' Check if all you need is installed
check_installed(pkgs = need_packages)

#load all packages, if needed
load_pkgs <-c("tibble", "readxl", "dplyr", "EnhancedVolcano", "ggplot2", "ggmanh", "extrafont","gap")
lapply(load_pkgs, require, character.only = TRUE )


```

### 4.1 Read in the data
#### 0.1 Directory variables

OBS: same paths on FEDORA
```{r paged.print=TRUE, root.dir=TRUE}
inputFolder <- "~/walssbasslab_server/Walss-Bass lab juliet/Iago/1.EWAS_K01_Analysis/Aug.24-EWAS_BDvsHC_BRAINBANK/2.Output_EWAS-models/4.DMPsSheets/1.Base-Model/"
outputFolder <- "~/walssbasslab_server/Walss-Bass lab juliet/Iago/1.EWAS_K01_Analysis/Aug.24-EWAS_BDvsHC_BRAINBANK/2.Output_EWAS-models"
knitr::opts_knit$set(root.dir = inputFolder)
```

OBS: Paths defined to the lab computer
```{r}
#OBS: Paths defined to the lab computer
#inputFolder <-  "V:/common/Walss-Bass_Lab/Walss-Bass lab juliet/Iago/1.EWAS_K01_Analysis/Aug.24-EWAS_BDNSvsBDS_BRAINBANK"
#outputFolder <- "~/walssbasslab_server/Walss-Bass lab juliet/Iago/1.EWAS_K01_Analysis/Aug.24-EWAS_BDNSvsBDS_BRAINBANK/2.Output_EWAS-models"
```


```{r}
# knitr::opts_knit$set(root.dir = inputFolder)
```


## 0.2 Input variables 
```{r}
date <- "Aug.2024"
# - BDNS vs. BDS DMPs - Accounting for smoking
DMPs <- read.csv(file.path(inputFolder,"EWAS-BB-BDSvsBDNS_DMPs_Aug.2024.csv"), row.names = 1)

# - BDNS vs. BDS DMPs svaPCs - Accounting for smoking
DMPsSVA <- read.csv(file.path(inputFolder,"EWAS-BB-BDSvsBDNS_DMPs_svaPCs_Aug.2024.csv"), row.names = 1)

```

## 0.3 Output variables 
### Manhatam
```{r}
BDNSvsBDS_ManhattanPlot <- paste0("EWAS-BB-BDNSvsBDS_ManhattanPlot_",date,".png")
BDNSvsBDS_svaPCs_ManhattanPlot <- paste0("EWAS-BB-BDNSvsBDS_svaPCs_ManhattanPlot_",date,".png") 
```

### QQplot 
```{r}
BDNSvsBDS_QQplot <- paste0("EWAS-BB-BDNSvsBDS_QQplot_",date,".png")
BDNSvsBDS_QQplot_svaPCs <- paste0("EWAS-BB-BDNSvsBDS_QQplot_svaPCs_",date,".png")
```
### Volcano Plot
```{r}
BDNSvsBDS_FDR_Corrected_VolcanoPlot <- paste0("EWAS-BB-BDNSvsBDS_FDR-Corrected_VolcanoPlot",date,".png")
BDNSvsBDS_FDR_Corrected_svaPCs_VolcanoPlot <- paste0("EWAS-BB-BDNSvsBDS_FDR-Corrected_svaPCs_VolcanoPlot",date,".png")
```


# BD vs HC Manhatam plot
### NO SVA 
```{r}

DMPs <- DMPs[which(DMPs$CHR %in% c(1:22)),]
DMPs$CHR <- as.numeric(DMPs$CHR)
DMPs_sorted <- DMPs[order(DMPs$CHR),]
#DMPs_sorted$CHR <- factor(DMPs_sorted$CHR) 


```
#### Defining thresholds
```{r}
# Define two threshold lines
EWSthresh <- -log10(9e-8)


# - Defining the FDR threshold line
p_values <- DMPs_sorted$P.Value

###############################################################################
# Function definition 
FDRthres_function <- function(PVAL,FDRLEVEL){
n <- length(PVAL)

# Sort p-values
sorted_PVAL <- sort(PVAL)

# Calculate BH critical values
bh_critical_values <- (1:n) / n * FDRLEVEL

# Find the largest p-value that is less than or equal to its BH critical value
threshold_index <- max(which(sorted_PVAL <= bh_critical_values))
fdr_threshold <- sorted_PVAL[threshold_index]
return(fdr_threshold)
}
###############################################################################
fdrthres <- FDRthres_function(DMPs_sorted$P.Value,0.05)
FDRthres <- -log10(fdrthres)
```
#### Generating Plot
```{r}

g <- DMPs_sorted %>% manhattan_plot(
  .,
  chromosome = NULL,
  signif = c(EWSthresh, FDRthres),
  pval.colname = "P.Value",
  chr.colname = "CHR",
  pos.colname = "MAPINFO",
  rescale = TRUE,
  annotatePval = 0.00000009)

g + 
  scale_color_manual(values = rep(c("gray1","gray"), length.out = 22)) +
  geom_hline(yintercept = EWSthresh, color = "red", linetype = "dashed") +
  geom_hline(yintercept = FDRthres, color = "blue4", linetype = "longdash") +
  ggtitle("BDNS vs. BDS Manhattan Plot") +
  theme(plot.title = element_text(size = 13, face = "bold", family = "Segoi U")) +
  ylim(0,10)
 
ggsave(filename = BDNSvsBDS_ManhattanPlot ,path = file.path(outputFolder,"2.ManhattanPlots"), device='png', dpi=300, width = 10, height = 7)

```
```{r}
# List of SNPs to highlight are in the snpsOfInterest object
# We will use ggrepel for the annotation
library(ggrepel)
DMPs_sorted <-DMPs_sorted[!is.na(DMPs_sorted),]
snpsOfInterest <- rownames(DMPs_sorted[which(DMPs_sorted$P.Value < 9e-8),])
# Prepare the dataset
don <- DMPs_sorted %>% 
  
  # Compute chromosome size
  group_by(CHR) %>% 
  summarise(chr_len=max(MAPINFO)) %>% 
  
  # Calculate cumulative position of each chromosome
  mutate(tot=cumsum(chr_len)-chr_len) %>%
  select(-chr_len) %>%
  
  # Add this info to the initial dataset
  left_join(DMPs_sorted, ., by=c("CHR"="CHR")) %>%
  
  # Add a cumulative position of each SNP
  arrange(CHR, MAPINFO) %>%
  mutate( Chromosome=MAPINFO+tot) %>%

  # Add highlight and annotation information
  mutate( is_highlight=ifelse(Name %in% snpsOfInterest, "yes", "no")) %>%
  mutate( is_annotate=ifelse(-log10(P.Value)>EWSthresh, "yes", "no")) 

# Prepare X axis
axisdf <- don %>% group_by(CHR) %>% summarize(center=( max(Chromosome) + min(Chromosome) ) / 2 )

# Make the plot
ggplot(don, aes(x=Chromosome, y=-log10(P.Value))) +
    
    # Show all points
    geom_point( aes(color=as.factor(CHR)), alpha=0.8, size=1.2) +
    scale_color_manual(values = rep(c("grey", "skyblue"), 22 )) +
    
    # custom X axis:
    scale_x_continuous( label = axisdf$CHR, breaks= axisdf$center ) +
    scale_y_continuous(expand = c(0, 0) ) +     # remove space between plot area and x axis

    # Add highlighted points
    geom_point(data=subset(don, is_highlight=="yes"), color="orange", size=2) +
  
    # Add label using ggrepel to avoid overlapping
    geom_label_repel( data=subset(don, is_annotate=="yes"), aes(label=Name), size=5) +

    # Custom the theme:
    theme_bw() +
    theme( 
      legend.position="none",
      panel.border = element_blank(),
      panel.grid.major.x = element_blank(),
      panel.grid.minor.x = element_blank()
    )
```

### With SVA 
```{r}

DMPsSVA <- DMPsSVA[which(DMPsSVA$CHR %in% c(1:22)),]
DMPsSVA$CHR <- as.numeric(DMPsSVA$CHR)
DMPsSVA_sorted <- DMPsSVA[order(DMPsSVA$CHR),]
#DMPsSVA_sorted$CHR <- factor(DMPsSVA_sorted$CHR) 


```
#### Defining thresholds
```{r}
# Define two threshold lines
EWSthresh <- -log10(9e-8)


# - Defining the FDR threshold line
p_values <- DMPsSVA_sorted$P.Value

###############################################################################
# Function definition 
FDRthres_function <- function(PVAL,FDRLEVEL){
n <- length(PVAL)

# Sort p-values
sorted_PVAL <- sort(PVAL)

# Calculate BH critical values
bh_critical_values <- (1:n) / n * FDRLEVEL

# Find the largest p-value that is less than or equal to its BH critical value
threshold_index <- max(which(sorted_PVAL <= bh_critical_values))
fdr_threshold <- sorted_PVAL[threshold_index]
return(fdr_threshold)
}
###############################################################################
fdrthres <- FDRthres_function(DMPsSVA_sorted$P.Value,0.05)
FDRthres <- -log10(fdrthres)
```
#### Generating Plot
```{r}

g <- DMPsSVA_sorted %>% manhattan_plot(
  .,
  chromosome = NULL,
  signif = c(EWSthresh, FDRthres),
  pval.colname = "P.Value",
  chr.colname = "CHR",
  pos.colname = "MAPINFO",
  rescale = FALSE)

g + 
  scale_color_manual(values = rep(c("gray1","gray"), length.out = 22)) +
  geom_hline(yintercept = EWSthresh, color = "red", linetype = "dashed") +
  geom_hline(yintercept = FDRthres, color = "blue4", linetype = "longdash") +
  ggtitle("BDNS vs. BDS Manhattan Plot - svaPCs  ") +
  theme(plot.title = element_text(size = 13, face = "bold", family = "Segoi U")) + 
  ylim(0,10)
ggsave(filename = BDNSvsBDS_svaPCs_ManhattanPlot,path = file.path(outputFolder,"2.ManhattanPlots/"), device='png', dpi=600, width = 10, height = 7)
```
# BDNS vs. BDS QQplot
### NO SVA
```{r}

lambda_gc <- gc.lambda(DMPs_sorted$P.Value)


qq <- ggmanh::qqunif(x= DMPs_sorted$P.Value)

qq + ggtitle("BDNS vs. BDS QQPlot - ") +
  theme(plot.title = element_text(size = 17, face = "bold", family = "Segoi U")) + 
  annotate("text", x = Inf, y = Inf, label = paste("lambda GC =", round(lambda_gc, 3)), 
           hjust = 1.1, vjust = 1.1, size = 5, color = "red")
ggsave(filename = BDNSvsBDS_QQplot,path = file.path(outputFolder,"1.QQplots/"), device='png', dpi=300, width = 7, height = 7)

```

### WITH SVA
```{r}

lambda_gc <- gc.lambda(DMPsSVA_sorted$P.Value)


qq <- ggmanh::qqunif(x= DMPsSVA_sorted$P.Value)

qq + ggtitle("BDNS vs. BDS QQPlot - svaPCs") +
  theme(plot.title = element_text(size = 15, face = "bold", family = "Segoi U")) + 
  annotate("text", x = Inf, y = Inf, label = paste("lambda GC =", round(lambda_gc, 3)), 
           hjust = 1.1, vjust = 1.1, size = 5, color = "red")
ggsave(filename = BDNSvsBDS_QQplot_svaPCs,path = file.path(outputFolder,"1.QQplots/"), device='png', dpi=300, width = 7, height = 7)
```

# BDNS vs. BDS Volcano plot
### NO SVA
Volcano plot with [EnhancedVolcano](https://bioconductor.org/packages/release/bioc/vignettes/EnhancedVolcano/inst/doc/EnhancedVolcano.html)
```{r}
################################################################################
# FDRcorrected p-value 
################################################################################
pv = 9e-8 # set your p-value
fc = 10^1000 # set your fold change

fdrthres <- FDRthres_function(DMPs$P.Value,0.05)

DMPs_volcano <- DMPs[,c(7,10)]

colnames(DMPs_volcano) = c('log2FoldChange', 'pvalue')

BDNSvsBDS_EnhVolc =EnhancedVolcano(DMPs_volcano,
  lab = NA,
  x = 'log2FoldChange',
  y = 'pvalue',
  pCutoff = pv,
  FCcutoff = fc,
  ylim = c(-0.5,10),
  xlim = c(-1.8,1.8),
  labSize = 2.0,
  hline = c(fdrthres,9e-8),
  hlineCol = c('red','blue'),
  legendLabels=c('Not sig.','log2 FC','EWS p-value'),
    legendPosition = 'right',
    legendLabSize = 16,
    legendIconSize = 5.0)
BDNSvsBDS_EnhVolc + ggtitle("BDNS vs. BDS Enhanced Volcano FDR-,Corrected p-value", 
                         subtitle = "") +
  theme(plot.title = element_text(size = 15, face = "bold", family = "Segoi U"))
#ggsave(paste0("VolcanoPlot_BD_v_HC_FDR-corrected",date,".png"),path = "V:/common/Walss-Bass_Lab/Walss-Bass lab juliet/Iago/1.EWAS_K01_Analysis/Aug.24-EWAS_BDNSvsBDS_BRAINBANK/2.Output_EWAS-models/3.VolcanoPlots", device='png', dpi=300, width = 10, height = 7)
ggsave(filename = BDNSvsBDS_FDR_Corrected_VolcanoPlot,path = file.path(outputFolder,"3.VolcanoPlots"), device='png', dpi=300, width = 10, height = 7)

```
```{r}
pv = EWSthresh
# set your p-value
fc = 0.5 # set your fold change

fdrthres <- FDRthres_function(DMPs$P.Value,0.05)
# Finally, we can organize the labels nicely using the "ggrepel" package and the geom_text_repel() function
# load library
# plot adding up all layers we have seen so far
BDNSvsBDS_ggplot = ggplot(data=DMPs_sorted, aes(x=logFC, y=-log10(P.Value))) +
        geom_point(aes(color=ifelse(-log10(P.Value) > -log10(9e-8), "above", "below")), show.legend = FALSE) + 
        geom_point(data=subset(DMPs_sorted, -log10(P.Value) > -log10(9e-8)), aes(x=logFC, y=-log10(P.Value)), color="orange", shape=18, size=3) +
        geom_text(data=subset(DMPs_sorted, -log10(P.Value) > -log10(9e-8)), aes(x=logFC, y=-log10(P.Value), label=Name), vjust=-1, size=3) +
        theme_linedraw() +
        scale_color_manual(values=c("above"="blue", "below"="azure4")) +
        geom_hline(aes(yintercept=-log10(9e-8), linetype="EWS p value"), col="deepskyblue2") +
        geom_hline(aes(yintercept=-log10(fdrthres), linetype="FDR corrected"), col="brown1") +
        scale_linetype_manual(name="Thresholds", values=c("EWS p value"="dashed", "FDR corrected"="dashed")) +
        guides(linetype=guide_legend(override.aes=list(color=c("deepskyblue2", "brown1")), title.position="top")) +
        ylim(-0.5,10)

BDNSvsBDS_ggplot + ggtitle("BDS vs. BDNS - Volcano Plot", 
                         subtitle = "") +
  theme(plot.title = element_text(size = 15, face = "bold", family = "Segoi U"))
ggsave(filename = BDNSvsBDS_FDR_Corrected_VolcanoPlot,path = file.path(outputFolder,"3.VolcanoPlots"), device='png', dpi=300, width = 10, height = 7)

```

### WITH SVA
```{r}
# ################################################################################
# # FDRcorrected p-value 
# ################################################################################
pv = EWSthresh # set your p-value
fc = 10^1000 # set your fold change
DMPsSVA_volcano <- DMPsSVA[,c(7,10)]

colnames(DMPsSVA_volcano) = c('log2FoldChange', 'pvalue')
BDNSvsBDS_EnhVolc =EnhancedVolcano(DMPsSVA_volcano,
  lab = NA,
  x = 'log2FoldChange',
  y = 'pvalue',
  pCutoff = pv,
  FCcutoff = fc,
  ylim = c(-0.5,10),
  xlim = c(-1.8,1.8),
  labSize = 2.0)
BDNSvsBDS_EnhVolc + ggtitle("BDNS vs. BDS Enhanced Volcano FDR-Corrected p-value", 
                         subtitle = "svaPCs ") +
  theme(plot.title = element_text(size = 15, face = "bold", family = "Segoi U"))
#ggsave(paste0("VolcanoPlot_BD_v_HC_FDR-corrected",date,".png"),path = "V:/common/Walss-Bass_Lab/Walss-Bass lab juliet/Iago/1.EWAS_K01_Analysis/Aug.24-EWAS_BDNSvsBDS_BRAINBANK/2.Output_EWAS-models/3.VolcanoPlots", device='png', dpi=300, width = 10, height = 7)
ggsave(filename = BDNSvsBDS_FDR_Corrected_svaPCs_VolcanoPlot,path = file.path(outputFolder,"3.VolcanoPlots"), device='png', dpi=300, width = 10, height = 7)

```
```{r}
fdrthres <- FDRthres_function(DMPsSVA$P.Value,0.05)
# Finally, we can organize the labels nicely using the "ggrepel" package and the geom_text_repel() function
# load library
# plot adding up all layers we have seen so far
BDNSvsBDS_ggplot = ggplot(data=DMPsSVA_sorted, aes(x=logFC, y=-log10(P.Value))) +
        geom_point(aes(color=ifelse(-log10(P.Value) > -log10(9e-8), "above", "below")), show.legend = FALSE) + 
        geom_point(data=subset(DMPsSVA_sorted, -log10(P.Value) > -log10(9e-8)), aes(x=logFC, y=-log10(P.Value)), color="orange", shape=18, size=3) +
        geom_text(data=subset(DMPsSVA_sorted, -log10(P.Value) > -log10(9e-8)), aes(x=logFC, y=-log10(P.Value), label=Name), vjust=-1, size=3) +
        theme_linedraw() +
        scale_color_manual(values=c("below"="azure4")) +
        geom_hline(aes(yintercept=-log10(9e-8), linetype="EWS p value"), col="deepskyblue2") +
        geom_hline(aes(yintercept=-log10(fdrthres), linetype="FDR corrected"), col="brown1") +
        scale_linetype_manual(name="Thresholds", values=c("EWS p value"="dashed", "FDR corrected"="dashed")) +
        guides(linetype=guide_legend(override.aes=list(color=c("deepskyblue2", "brown1")), title.position="top")) +
        ylim(-0.5,10)

BDNSvsBDS_ggplot + ggtitle("BDS vs. BDNS - Volcano Plot", 
                         subtitle = "svaPCs - sensitivity analysis") +
  theme(plot.title = element_text(size = 15, face = "bold", family = "Segoi U"))
ggsave(filename = BDNSvsBDS_FDR_Corrected_svaPCs_VolcanoPlot,path = file.path(outputFolder,"3.VolcanoPlots"), device='png', dpi=300, width = 10, height = 7)
```

```{r}
print("END OF SCRIPT")
```

