---
title: "BDNS vs HC "
author: "Iago"
date: "`r Sys.Date()`"
output: html_document
---


```{r, results='hide'}
source("~/walssbasslab_server/Walss-Bass lab juliet/Iago/1.EWAS_K01_Analysis/Aug.24-EWAS_BDvsHC_BRAINBANK/source_functions/source_functions.R")

# windows path
#source("V:/common/Walss-Bass_Lab/Walss-Bass lab juliet/Iago/1.EWAS_K01_Analysis/Aug.24-EWAS_BDNSvsHC_BRAINBANK/source_functions/source_functions.R")

need_packages <-c("tibble", "readxl", "dplyr", "EnhancedVolcano", "ggplot2", "ggmanh", "extrafont","gap")#' install by running the function
install_cran_pkgs(pkgs = need_packages)
install_bioconductor_pkgs(pkgs = need_packages)

#' Check if all you need is installed
check_installed(pkgs = need_packages)

#load all packages, if needed
load_pkgs <-c("tibble", "readxl", "dplyr", "EnhancedVolcano", "ggplot2", "ggmanh", "extrafont","gap")
lapply(load_pkgs, require, character.only = TRUE )


```

### 4.1 Read in the data
#### 0.1 Directory variables

OBS: same paths on FEDORA
```{r paged.print=TRUE, root.dir=TRUE}
inputFolder <- "~/walssbasslab_server/Walss-Bass lab juliet/Iago/1.EWAS_K01_Analysis/Aug.24-EWAS_BDvsHC_BRAINBANK/2.Output_EWAS-models/4.DMPsSheets/"
outputFolder <- "~/walssbasslab_server/Walss-Bass lab juliet/Iago/1.EWAS_K01_Analysis/Aug.24-EWAS_BDvsHC_BRAINBANK/2.Output_EWAS-models"
knitr::opts_knit$set(root.dir = inputFolder)
```

OBS: Paths defined to the lab computer
```{r}

#OBS: Paths defined to the lab computer
# #inputFolder <-  "V:/common/Walss-Bass_Lab/Walss-Bass lab juliet/Iago/1.EWAS_K01_Analysis/Aug.24-EWAS_BDNSvsHC_BRAINBANK"
# outputFolder <- "~/walssbasslab_server/Walss-Bass lab juliet/Iago/1.EWAS_K01_Analysis/Aug.24-EWAS_BDNSvsHC_BRAINBANK/2.Output_EWAS-models/"
# knitr::opts_knit$set(root.dir = inputFolder)
```


## 0.2 Input variables 
```{r}
date <- "Aug.2024"
# - BDNS vs. HC DMPs
DMPs <- read.csv(file.path(inputFolder,"1.Base-Model/EWAS-BB-BDNSvsHC_DMPs_Aug.2024.csv"), row.names = 1)

# - BDNS vs. HC DMPs svaPCs
DMPsSVA <- read.csv(file.path(inputFolder,"1.Base-Model/EWAS-BB-BDNSvsHC_DMPs_svaPCs_Aug.2024.csv"), row.names = 1)

```

## 0.3 Output variables 
### Manhatam
```{r}
BDNSvsHC_ManhattanPlot <- paste0("EWAS-BB-BDNSvsHC_ManhattanPlot_",date,".png")
BDNSvsHC_svaPCs_ManhattanPlot <- paste0("EWAS-BB-BDNSvsHC_svaPCs_ManhattanPlot_",date,".png") 
```

### QQplot 
```{r}
BDNSvsHC_QQplot <- paste0("EWAS-BB-BDNSvsHC_QQplot_",date,".png")
BDNSvsHC_QQplot_svaPCs <- paste0("EWAS-BB-BDNSvsHC_QQplot_svaPCs_",date,".png")
```
### Volcano Plot
```{r}
BDNSvsHC_FDR_Corrected_VolcanoPlot <- paste0("EWAS-BB-BDNSvsHC_FDR-Corrected_VolcanoPlot",date,".png")
BDNSvsHC_FDR_Corrected_svaPCs_VolcanoPlot <- paste0("EWAS-BB-BDNSvsHC_FDR-Corrected_svaPCs_VolcanoPlot",date,".png")
```


# BD vs HC Manhatam plot
### NO SVA 
```{r}

DMPs <- DMPs[which(DMPs$CHR %in% c(1:22)),]
DMPs$CHR <- as.numeric(DMPs$CHR)
DMPs_sorted <- DMPs[order(DMPs$CHR),]
#DMPs_sorted$CHR <- factor(DMPs_sorted$CHR) 


```
#### Defining thresholds
```{r}
# Define two threshold lines
EWSthresh <- -log10(9e-8)


# - Defining the FDR threshold line
p_values <- DMPs_sorted$P.Value

###############################################################################
# Function definition 
FDRthres_function <- function(PVAL,FDRLEVEL){
n <- length(PVAL)

# Sort p-values
sorted_PVAL <- sort(PVAL)

# Calculate BH critical values
bh_critical_values <- (1:n) / n * FDRLEVEL

# Find the largest p-value that is less than or equal to its BH critical value
threshold_index <- max(which(sorted_PVAL <= bh_critical_values))
fdr_threshold <- sorted_PVAL[threshold_index]
return(fdr_threshold)
}
###############################################################################
fdrthres <- FDRthres_function(DMPs_sorted$P.Value,0.05)
FDRthres <- -log10(fdrthres)
```
#### Generating Plot
```{r}

g <- DMPs_sorted %>% manhattan_plot(
  .,
  chromosome = NULL,
  signif = c(EWSthresh, FDRthres),
  pval.colname = "P.Value",
  chr.colname = "CHR",
  pos.colname = "MAPINFO",
  rescale = TRUE)

g + 
  scale_color_manual(values = rep(c("gray1","gray"), length.out = 22)) +
  geom_hline(yintercept = EWSthresh, color = "red", linetype = "dashed") +
  geom_hline(yintercept = FDRthres, color = "blue4", linetype = "longdash") +
  ggtitle("BDNS vs. HC Manhattan Plot") +
  theme(plot.title = element_text(size = 15, face = "bold", family = "Segoi U")) +
  ylim(0, 10) # Adjust max_y_value to your desired upper limit
ggsave(filename = BDNSvsHC_ManhattanPlot ,path = file.path(outputFolder,"2.ManhattanPlots/"), device='png', dpi=300, width = 10, height = 7)

```
### With SVA 
```{r}

DMPsSVA <- DMPsSVA[which(DMPsSVA$CHR %in% c(1:22)),]
DMPsSVA$CHR <- as.numeric(DMPsSVA$CHR)
DMPsSVA_sorted <- DMPsSVA[order(DMPsSVA$CHR),]
#DMPsSVA_sorted$CHR <- factor(DMPsSVA_sorted$CHR) 


```
#### Defining thresholds
```{r}
# Define two threshold lines
EWSthresh <- -log10(9e-8)


# - Defining the FDR threshold line
p_values <- DMPsSVA_sorted$P.Value

###############################################################################
# Function definition 
FDRthres_function <- function(PVAL,FDRLEVEL){
n <- length(PVAL)

# Sort p-values
sorted_PVAL <- sort(PVAL)

# Calculate BH critical values
bh_critical_values <- (1:n) / n * FDRLEVEL

# Find the largest p-value that is less than or equal to its BH critical value
threshold_index <- max(which(sorted_PVAL <= bh_critical_values))
fdr_threshold <- sorted_PVAL[threshold_index]
return(fdr_threshold)
}
###############################################################################
fdrthres <- FDRthres_function(DMPsSVA_sorted$P.Value,0.05)
FDRthres <- -log10(fdrthres)
```
#### Generating Plot
```{r}

g <- DMPsSVA_sorted %>% manhattan_plot(
  .,
  chromosome = NULL,
  signif = c(EWSthresh, FDRthres),
  pval.colname = "P.Value",
  chr.colname = "CHR",
  pos.colname = "MAPINFO",
  rescale = TRUE)

g + 
  scale_color_manual(values = rep(c("gray1","gray"), length.out = 22)) +
  geom_hline(yintercept = EWSthresh, color = "red", linetype = "dashed") +
  geom_hline(yintercept = FDRthres, color = "blue4", linetype = "longdash") +
  ggtitle("BDNS vs. HC Manhattan Plot - svaPCs") +
  theme(plot.title = element_text(size = 12, face = "bold", family = "Segoi U"))+
  ylim(0, 10) # Adjust max_y_value to your desired upper limit

ggsave(filename = BDNSvsHC_svaPCs_ManhattanPlot,path = file.path(outputFolder,"2.ManhattanPlots/"), device='png', dpi=300, width = 10, height = 7)
```
# BDNS vs. HC QQplot
### NO SVA
```{r}

lambda_gc <- gc.lambda(DMPs_sorted$P.Value)


qq <- ggmanh::qqunif(x= DMPs_sorted$P.Value)

qq + ggtitle("BDNS vs. HC QQPlot") +
  theme(plot.title = element_text(size = 17, face = "bold", family = "Segoi U")) + 
  annotate("text", x = Inf, y = Inf, label = paste("lambda GC =", round(lambda_gc, 3)), 
           hjust = 1.1, vjust = 1.1, size = 5, color = "red")
ggsave(filename = BDNSvsHC_QQplot,path = file.path(outputFolder,"1.QQplots/"), device='png', dpi=300, width = 7, height = 7)

```

### WITH SVA
```{r}

lambda_gc <- gc.lambda(DMPsSVA_sorted$P.Value)


qq <- ggmanh::qqunif(x= DMPsSVA_sorted$P.Value)

qq + ggtitle("BDNS vs. HC QQPlot - svaPCs") +
  theme(plot.title = element_text(size = 15, face = "bold", family = "Segoi U")) + 
  annotate("text", x = Inf, y = Inf, label = paste("lambda GC =", round(lambda_gc, 3)), 
           hjust = 1.1, vjust = 1.1, size = 5, color = "red")
ggsave(filename = BDNSvsHC_QQplot_svaPCs,path = file.path(outputFolder,"1.QQplots/"), device='png', dpi=300, width = 7, height = 7)
```

# BDNS vs. HC Volcano plot
### NO SVA
Volcano plot with [EnhancedVolcano](https://bioconductor.org/packages/release/bioc/vignettes/EnhancedVolcano/inst/doc/EnhancedVolcano.html)
```{r}
################################################################################
# 0.05 nominal threshold
################################################################################
# pv = 0.05 # set your p-value
# fc = 0.5 # set your fold change
# 
# DMPs_volcano <- DMPs[,c(7,10)]
# 
# colnames(DMPs_volcano) = c('log2FoldChange', 'pvalue')
# library(EnhancedVolcano)
# BDNSvsHC_EnhVolc =EnhancedVolcano(DMPs_volcano,
#   lab = NA,
#   x = 'log2FoldChange',
#   y = 'pvalue',
#   pCutoff = pv,
#   FCcutoff = fc,
#   ylim = c(-0.5,10),
#   xlim = c(-1.8,1.8),
#   labSize = 2.0,)
# 
# BDNSvsHC_EnhVolc + ggtitle("BDNS vs. HC Enhanced Volcano Nominal p-value",
#                          subtitle = "") +
#   theme(plot.title = element_text(size = 15, face = "bold", family = "Segoi U"))
# #ggsave(paste0("VolcanoPlot_BD_v_HC_Nominal-p",date,".png"),path = "V:/common/Walss-Bass_Lab/Walss-Bass lab juliet/Iago/1.EWAS_K01_Analysis/Aug.24-EWAS_BDNSvsHC_BRAINBANK/2.Output_EWAS-models/3.VolcanoPlots", device='png', dpi=300, width = 10, height = 7)
# ggsave(paste0("EWAS-BB-BDNSvsHC_Nominal-p_VolcanoPlot",date,".png"),path = file.path(outputFolder,"3.VolcanoPlots"), device='png', dpi=300, width = 10, height = 7)
################################################################################
# FDRcorrected p-value 
################################################################################
pv = fdrthres # set your p-value
fc = 10^1000 # set your fold change
DMPs_volcano <- DMPs[,c(7,10)]

colnames(DMPs_volcano) = c('log2FoldChange', 'pvalue')
library(EnhancedVolcano)
BDNSvsHC_EnhVolc =EnhancedVolcano(DMPs_volcano,
  lab = NA,
  x = 'log2FoldChange',
  y = 'pvalue',
  pCutoff = pv,
  FCcutoff = fc,
  ylim = c(-0.5,10),
  xlim = c(-1.8,1.8),
  labSize = 2.0)
BDNSvsHC_EnhVolc + ggtitle("BDNS vs. HC Enhanced Volcano FDR-Corrected p-value", 
                         subtitle = "") +
  theme(plot.title = element_text(size = 15, face = "bold", family = "Segoi U"))
#ggsave(paste0("VolcanoPlot_BD_v_HC_FDR-corrected",date,".png"),path = "V:/common/Walss-Bass_Lab/Walss-Bass lab juliet/Iago/1.EWAS_K01_Analysis/Aug.24-EWAS_BDNSvsHC_BRAINBANK/2.Output_EWAS-models/3.VolcanoPlots", device='png', dpi=300, width = 10, height = 7)
ggsave(filename = BDNSvsHC_FDR_Corrected_VolcanoPlot,path = file.path(outputFolder,"3.VolcanoPlots"), device='png', dpi=300, width = 10, height = 7)

```
```{r}
pv = EWSthresh
# set your p-value
fc = 0.5 # set your fold change

fdrthres <- FDRthres_function(DMPs$P.Value,0.05)
# Finally, we can organize the labels nicely using the "ggrepel" package and the geom_text_repel() function
# load library
# plot adding up all layers we have seen so far
BDNSvsHC_ggplot = ggplot(data=DMPs_sorted, aes(x=logFC, y=-log10(P.Value))) +
        geom_point(aes(color=ifelse(-log10(P.Value) > -log10(9e-8), "above", "below")), show.legend = FALSE) + 
        geom_point(data=subset(DMPs_sorted, -log10(P.Value) > -log10(9e-8)), aes(x=logFC, y=-log10(P.Value)), color="orange", shape=18, size=3) +
        geom_text(data=subset(DMPs_sorted, -log10(P.Value) > -log10(9e-8)), aes(x=logFC, y=-log10(P.Value), label=Name), vjust=-1, size=3) +
        theme_linedraw() +
        scale_color_manual(values=c("above"="blue", "below"="azure4")) +
        geom_hline(aes(yintercept=-log10(9e-8), linetype="EWS p value"), col="deepskyblue2") +
        geom_hline(aes(yintercept=-log10(fdrthres), linetype="FDR corrected"), col="brown1") +
        scale_linetype_manual(name="Thresholds", values=c("EWS p value"="dashed", "FDR corrected"="dashed")) +
        guides(linetype=guide_legend(override.aes=list(color=c("deepskyblue2", "brown1")), title.position="top")) +
        ylim(-0.5,10)

BDNSvsHC_ggplot + ggtitle("BDNS vs. HC - Volcano Plot", 
                         subtitle = "") +
  theme(plot.title = element_text(size = 15, face = "bold", family = "Segoi U"))
ggsave(filename = BDNSvsHC_FDR_Corrected_VolcanoPlot,path = file.path(outputFolder,"3.VolcanoPlots"), device='png', dpi=300, width = 10, height = 7)
```

### WITH SVA
```{r}

# 
# ################################################################################
# # FDRcorrected p-value 
# ################################################################################
pv = fdrthres # set your p-value
fc = 10^1000 # set your fold change
DMPsSVA_volcano <- DMPsSVA[,c(7,10)]

colnames(DMPsSVA_volcano) = c('log2FoldChange', 'pvalue')
library(EnhancedVolcano)
BDNSvsHC_EnhVolc =EnhancedVolcano(DMPsSVA_volcano,
  lab = NA,
  x = 'log2FoldChange',
  y = 'pvalue',
  pCutoff = pv,
  FCcutoff = fc,
  ylim = c(-0.5,10),
  xlim = c(-1.8,1.8),
  labSize = 2.0)
BDNSvsHC_EnhVolc + ggtitle("BDNS vs. HC Enhanced Volcano FDR-Corrected p-value", 
                         subtitle = "svaPCs and ") +
  theme(plot.title = element_text(size = 15, face = "bold", family = "Segoi U"))
#ggsave(paste0("VolcanoPlot_BD_v_HC_FDR-corrected",date,".png"),path = "V:/common/Walss-Bass_Lab/Walss-Bass lab juliet/Iago/1.EWAS_K01_Analysis/Aug.24-EWAS_BDNSvsHC_BRAINBANK/2.Output_EWAS-models/3.VolcanoPlots", device='png', dpi=300, width = 10, height = 7)
ggsave(filename = BDNSvsHC_FDR_Corrected_svaPCs_VolcanoPlot,path = file.path(outputFolder,"3.VolcanoPlots"), device='png', dpi=300, width = 10, height = 7)

```
```{r}
fdrthres <- FDRthres_function(DMPsSVA$P.Value,0.05)
# Finally, we can organize the labels nicely using the "ggrepel" package and the geom_text_repel() function
# load library
# plot adding up all layers we have seen so far
BDNSvsHC_ggplot = ggplot(data=DMPsSVA_sorted, aes(x=logFC, y=-log10(P.Value))) +
        geom_point(aes(color=ifelse(-log10(P.Value) > -log10(9e-8), "above", "below")), show.legend = FALSE) + 
        geom_point(data=subset(DMPsSVA_sorted, -log10(P.Value) > -log10(9e-8)), aes(x=logFC, y=-log10(P.Value)), color="orange", shape=18, size=3) +
        geom_text(data=subset(DMPsSVA_sorted, -log10(P.Value) > -log10(9e-8)), aes(x=logFC, y=-log10(P.Value), label=Name), vjust=-1, size=3) +
        theme_linedraw() +
        scale_color_manual(values=c("below"="azure4")) +
        geom_hline(aes(yintercept=-log10(9e-8), linetype="EWS p value"), col="deepskyblue2") +
        geom_hline(aes(yintercept=-log10(fdrthres), linetype="FDR corrected"), col="brown1") +
        scale_linetype_manual(name="Thresholds", values=c("EWS p value"="dashed", "FDR corrected"="dashed")) +
        guides(linetype=guide_legend(override.aes=list(color=c("deepskyblue2", "brown1")), title.position="top")) +
        ylim(-0.5,10)

BDNSvsHC_ggplot + ggtitle("BDNS vs. HC - Volcano Plot", 
                         subtitle = "svaPCs - sensitivity analysis") +
  theme(plot.title = element_text(size = 15, face = "bold", family = "Segoi U"))
ggsave(filename = BDNSvsHC_FDR_Corrected_svaPCs_VolcanoPlot,path = file.path(outputFolder,"3.VolcanoPlots"), device='png', dpi=300, width = 10, height = 7)
```

```{r}

```

