---
title: "EWAS-MinfiQC 1 - Data cleaning, RGset generation and minfi intensity QC metrics" 
author: "Iago"
date: "2024-09-17"
output:
  pdf_document: default
  html_notebook: default
---
This script is dedicated to realize minfi 1st step of quality control.

Stage Description:  
  1.) At first, this script will check the existence of all .idat files for all individuals. 
      Then, create RGset, Mset and GMset raw objects (Raw = Non-Normalized).  
  
  2.) Using GMRset, ccheck for sex discrepancy between sample.  
    a) Removing sex discrepant samples  
  
  3.) Minfi uMed/mMed quality control (Intensity QC metrics)  
  


```{r message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
source("~/walssbasslab_server/Walss-Bass lab juliet/Iago/1.EWAS_K01_Analysis/Aug.24-EWAS_BDvsHC_BRAINBANK/source_functions/source_functions.R")

# windows path
#source("V:/common/Walss-Bass_Lab/Walss-Bass lab juliet/Iago/1.EWAS_K01_Analysis/Aug.24-EWAS_BDvsHC_BRAINBANK/source_functions/source_functions.R")

need_packages <-c("tibble", "readxl","minfi", "readxl", "stringr", "dplyr", "IlluminaHumanMethylationEPICmanifest", "wateRmelon", "ewastools")#' install by running the function
install_cran_pkgs(pkgs = need_packages)
install_bioconductor_pkgs(pkgs = need_packages)

#' Check if all you need is installed
check_installed(pkgs = need_packages)

#load all packages, if needed
load_pkgs <- c("minfi", "readxl", "stringr", "dplyr")
lapply(load_pkgs, require, character.only = TRUE )


```

#### 0.1 Directory variables ----
```{r}
IdatFolder <-  "/home/frieslab/walssbasslab_server/Walss-Bass lab juliet/Gabriel/K01 analyses/iScan runs K01/idat_K01"
outputFolder <- "/home/frieslab/walssbasslab_server/Walss-Bass lab juliet/Iago/1.EWAS_K01_Analysis/Aug.24-EWAS_BDvsHC_BRAINBANK"
knitr::opts_chunk$set(echo = TRUE)
```

#### 0.2 Inut variables ----
```{r}
date <- "Aug.2024"
input_file <-  file.path(outputFolder,"4.Original_MetaData/K01_annotation_7.25.23_FullMetaData.xlsx")
```

#### 0.3 Output variables ----
```{r}
QC_metrics_file <- file.path(outputFolder,"7.Intermediate_files/QC_metrics_BRAIN-BANK.csv")
scatter_out_file <- file.path(outputFolder,"8.Plots/BRAIN-BANK_scatter_plot.png" )
```


# 1) Data imputation ----

#### 1.1) Input phenotype table
```{r Inputing Basenames paths to idat files} 
targets <- as.data.frame(read_excel(input_file))

if ("Basename" %in% colnames(targets)) {
  cat("The 'Basename' column exists in the dataframe.\n")
} else {
	targets$Basename <- paste0(IdatFolder,"/",targets$Sentrix_ID,"_",targets$Sentrix_Position)

  cat("The 'Basename' column does not exist in the dataframe.\n")
}
```

```{r Creating "Barcode" column based on the idat files present in the folder}
fileName_list <- list.files(path = IdatFolder, pattern = "idat")
barcode <- str_split(fileName_list, "_")

```

```{r Checking for duplicated barcode names in the folder}
inFolder_barcode <- list()
for(i in seq(length(barcode)) ){
  inFolder_barcode <- c(inFolder_barcode, paste0(barcode[[i]][1],"_",barcode[[i]][2]) )
}

head(inFolder_barcode)

inFolder_barcode_nodup <- inFolder_barcode[which(duplicated(inFolder_barcode))]
```
```{r Subsetting targets to only the ones we actually have data }
targetsInFolder <- subset(targets, targets$barcode %in% inFolder_barcode_nodup)

targets <- targetsInFolder

```



#### 1.3) RGset and Mset geneation  
--- Red & Green channel set and Methylation set --- 
```{r Creating RGset based on individuals showed in targets (metadata) }
cat("1.3.1 - Reading metharray to RGset")
if (file.exists(file.path(outputFolder,"7.Intermediate_files/RGset_BRAIN-BANK.RData"))){
  
  
  load(file.path(outputFolder,"7.Intermediate_files/RGset_BRAIN-BANK.RData"))

  } else{

    RGset <- read.metharray.exp(targets = targets, extended =TRUE, recursive = TRUE, verbose=TRUE, force=TRUE)

    save(RGset, file = file.path(outputFolder,"7.Intermediate_files/RGset_BRAIN-BANK.RData"))
}
```

#### 1.3.1 Creating MSet and GMset
  Obs: If you want to make the script faster, remove the saving step from MSet 
  and GMSset, its not necessary to have them being saved
```{r Creating MSet based on RGset}

cat("1.3.2 - Generating MSet, GMSet and Sex Check \n")
if(file.exists(file.path(outputFolder,"/7.Intermediate_files/MSet_BRAIN-BANK.RData"))){
  load(file.path(outputFolder,"/7.Intermediate_files/MSet_BRAIN-BANK.RData"))
} else {

  MSet <- preprocessRaw(RGset)

  print(MSet)

  save(MSet, file = file.path(outputFolder,"/7.Intermediate_files/MSet_BRAIN-BANK.RData"))
}


```
### GMSset
```{r Creating GMSet baed on RGset}

if(file.exists(file.path(outputFolder,"/7.Intermediate_files/GMset_BRAIN-BANK.RData"))){
  load(file.path(outputFolder,"/7.Intermediate_files/GMset_BRAIN-BANK.RData"))
} else {
GMset <- mapToGenome(MSet)
print(GMset)
save(GMset, file = file.path(outputFolder,"/7.Intermediate_files/GMset_BRAIN-BANK.RData"))
}

```

### GetSex() ----

```{r Checking sex discrepancy}


predSex <- getSex(GMset) %>% as.data.frame() ; predSex$barcode <- row.names(predSex)
head(predSex)


targets <- inner_join(x=targets, y = predSex, by="barcode")

#View(targets)


targets$Sex <- ifelse(targets$Sex %in% c("Male", "M"), "M", "F")

if(!all(targets$Sex == targets$predictedSex)){
    print("There are sex discordance in te sample")
    mismatch <- targets[which(targets$Sex != targets$predictedSex),]
    print(mismatch[,c("barcode","Sex","predictedSex")])
    write.csv(mismatch, file = file.path(outputFolder,paste0("6.Phenotype_files/K01_FailedSexMismatched_file",date,".csv")), row.names = FALSE, quote = TRUE)
} else{
    print("There ara no sex discordande in the sample")
}


targets$SexMismatch <- ifelse(targets$Sex == targets$predictedSex, FALSE, TRUE )


```


#### Minfi intensity quality control ----
```{r}
cat("1.3.3 - MSet quality control")

qc <- getQC(MSet)
head(qc)
QCmetrics <- cbind(targets,qc)



```


### 1.4) Writting mMed/uMed Quality Checks metrics and Generating meth/unmeth scatter plot  
--- Quality check graph ---  

**Question: what do you think about those below threshold? Should we remove?**

```{r}

QCmetrics$minfiFailed <- ifelse((QCmetrics$mMed + QCmetrics$uMed)/2 < 10.5, TRUE, FALSE)
print(table(QCmetrics$minfiFailed, QCmetrics$SexMismatch))

write.csv(QCmetrics, file=QC_metrics_file,row.names=FALSE)
plotQC(qc)
```

```{r}
rm(list = ls())
```