---
title: " EWAS minfi QC 2 - Bisulphite convertion and DetectionP"
output: html_notebook
---
Stage Description:  

  2.1) Remove probes which bisulphite convertion is below 80%  
  
  2.2) Read in the list of probes that are cross-hybridizing and are located  
  in SNPS  
  
  2.3) DetP value threshold = 0.001  
    a) Remove samples which detP mean is > 0.001  
    b) Remove probes which detP mean is > 0.001  
  
  2.4) Remove probes located in SNPs  
  
  2.5) Create a new RGset with bad samples and probes removed
```{r message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
source("~/walssbasslab_server/Walss-Bass lab juliet/Iago/1.EWAS_K01_Analysis/Aug.24-EWAS_BDvsHC_BRAINBANK/source_functions/source_functions.R")

# windows path
#source("V:/common/Walss-Bass_Lab/Walss-Bass lab juliet/Iago/1.EWAS_K01_Analysis/Aug.24-EWAS_BDvsHC_BRAINBANK/source_functions/source_functions.R")

need_packages <-c("tibble", "readxl","minfi", "readxl", "stringr", "dplyr", "IlluminaHumanMethylationEPICmanifest", "wateRmelon", "ewastools")#' install by running the function
install_cran_pkgs(pkgs = need_packages)
install_bioconductor_pkgs(pkgs = need_packages)

#' Check if all you need is installed
check_installed(pkgs = need_packages)

#load all packages, if needed
load_pkgs <- c("minfi", "readxl", "stringr", "dplyr","IlluminaHumanMethylationEPICmanifest","ewastools")
lapply(load_pkgs, require, character.only = TRUE )
```


#### 0.1 Directory variables
```{r}
inputFolder <-"/home/frieslab/walssbasslab_server/Walss-Bass lab juliet/Iago/1.EWAS_K01_Analysis/Aug.24-EWAS_BDvsHC_BRAINBANK"
outputFolder <- inputFolder
setwd(outputFolder)
```


#### 0.2 Inut variables 
```{r}
date <- "Aug.2024"
input_file <- file.path(inputFolder, "7.Intermediate_files/QC_metrics_BRAIN-BANK.csv")
remove <- file.path(outputFolder,"cross_reactiveEPIC_Iago.csv")
targets <- read.csv(file.path(outputFolder,"7.Intermediate_files/QC_metrics_BRAIN-BANK.csv"))
```

## 2.0 Loading RGset again  
--- Illumina Backgrouond correction ---  

```{r}
if(file.exists(file.path(inputFolder,paste0("7.Intermediate_files/RGset_bgcorr_BRAIN-BANK_",date,".RData")))) {
  load(file.path(inputFolder, paste0("7.Intermediate_files/RGset_bgcorr_BRAIN-BANK_",date,".RData"))) 
} else{
  RGset <- bgcorrect.illumina(RGset)  
  save(RGset, file= file.path(inputFolder,paste0("7.Intermediate_files/RGset_bgcorr_BRAIN-BANK_",date,".RData"))) 
}
```


## 2.1 Bisulphite convertion filtering  
--- Bisulphite convertion threshold > 80% ---  
  2.2.1 Creating a list of individuals that are 80% higher Bisulphite converstion 

```{r}
# # Make a list of targets that failed in minfiQC and SexCheck
# targets <- read.csv(input_file, stringsAsFactors = TRUE, header = TRUE)
# filtered_targets<- targets[!(targets$SexMismatch | targets$minfiFailed), ]
```

```{r}

Bisulphite <-bscon(RGset)

print("#make a list of individuals whose bisulphite conversion is >80%")
good.samples <- Bisulphite>80
good.samples <- good.samples[good.samples==TRUE] 
good.samples <- names(good.samples)  

```


## 2.2) Cross-hybridizing probes and SNP-prove removing  
```{r}
print("Reading cross-hybridizing file")
probes=read.table(remove,h=F)

print("Preliminary filtering by detection p-value")

if(file.exists(file.path(inputFolder,"7.Intermediate_files/detP_BRAIN-BANK.RData"))){
  load(file.path(inputFolder,"7.Intermediate_files/detP_BRAIN-BANK.RData"))
} else{
  detP <- minfi::detectionP(RGset)
save(detP, file = file.path(inputFolder,"7.Intermediate_files/detP_BRAIN-BANK.RData"))
}

```

## 2.3 Detection P value filtering step
```{r}
keep <- colMeans(detP) < 0.001 
failedProbes <- rownames(detP)[rowMeans(detP)>0.001] 
```
```{r}
barplot(colMeans(detP), las=2, cex.names=0.8, ylab="Mean detection p-values")
abline(h=0.001,col="red")
```


### 2.3.1 Removing the cross-hybridizin probes, bad detP samples and probes located in SNPs
```{r}
print("Creating a new RGset with bad samples and probes removed")
RGset <- subsetByLoci(rgSet=RGset[,keep],excludeLoci = c(probes,failedProbes), keepSnps = FALSE)
RGset <- RGset[ ,intersect(good.samples, colnames(RGset))]

save(RGset, file = file.path(outputFolder, paste0("7.Intermediate_files/RGset_goodsamples_no-SNPs_BRAIN-BANK_",date,".RData")))
```

```{r}

print("Removing bad samples from targets datasheet")
targets2 <- targets[which(targets$barcode %in% colnames(RGset) ),]

head(targets2)
targets <- targets2
dim(targets)
write.csv(targets, file = file.path(outputFolder, paste0("6.Phenotype_files/K01_annotation_QCmetrics_removal",date,".csv")))
```

```{r}

detP_good <- detP[!(rownames(detP) %in% failedProbes), which(colnames(detP) %in% names( keep[which(keep ==TRUE)] ) ) ]
ncol(detP_good)
```
```{r}
barplot(colMeans(detP_good), las=2, cex.names=0.8, ylab="Mean detection p-values")
abline(h=0.01,col="red")
```

```{r}
rm(list=ls())